Male cuticular pheromones stimulate removal of the mating plug and promote re-mating through pC1 neurons in Drosophila females

Version 1

Abstract

Abstract

In birds and insects, females uptake sperm for a specific duration post-copulation known as the ejaculate holding period (EHP) before expelling unused sperm and the mating plug through sperm ejection. Our study uncovered that encountering males or mated females after mating substantially shortens EHP, a phenomenon we term ‘male-induced EHP shortening (MIES)’. MIES requires Or47b+ olfactory and ppk23+ gustatory neurons, activated by 2-methyltetracosane and 7-Tricosene, respectively. These odorants raise cAMP levels in pC1 neurons, responsible for processing male courtship and regulating female mating receptivity. Elevated cAMP levels in pC1 neurons reduce EHP and reinstate their responsiveness to male courtship cues, promoting re-mating with faster sperm ejection. This study establishes MIES as a genetically tractable model of sexual plasticity with a conserved neural mechanism.

Introduction

Sexual plasticity, the ability to modify sexual state or reproductive behavior in response to changing social conditions, is observed in both vertebrates and invertebrates (1–5). In rodents, exposure to unfamiliar males often leads to the sudden termination of pregnancy, known as the Bruce effect. It is induced by male urinary peptides, such as MHC I peptides, activating the vomeronasal organ (6–8). This effect enhances reproductive fitness of both sexes, by eliminating the offspring of competing males and enabling females to select better mates even after conception. Many species also adapt their reproductive behaviors in response to the social sexual context change (SSCC), involving encounters with new sexual partners or competitors. Understanding the neuronal circuit mechanisms behind female responses to SSCC emerges as a central focus in neuroscience (9–12).

Drosophila melanogaster, the fruit fly, displays various social behaviors like aggregation, aggression, and sexual behavior (13–15). Similar to rodents, they primarily use the olfactory system to communicate socially through pheromones (16, 17). Some of these pheromones act as aphrodisiacs, while others regulate aggression or foster aggregation. For instance, cis-vaccenyl acetate (cVA) attracts females but repels males and promotes aggregation (14, 18, 19). 7-Tricosene (7-T), a cuticular hydrocarbon (CHC) present in males, is an aphrodisiac to females and affects social interactions between males (20, 21). On the other hand, 7,11-Heptacosadiene (7,11-HD), a related female-specific pheromone, functions as an aphrodisiac to males, triggering courtship behavior and involving species recognition (22, 23).

The fruit fly’s chemo-sensory organs, present in various body parts, detect these pheromones (24, 25). Olfactory receptor neurons (ORNs) in sensilla of the antenna and maxillary palps pick up long-range volatile pheromones like cVA, while short-range pheromones like 7-T are sensed by neurons on the forelegs and labellum (16, 17, 24).

The olfactory receptor Or47b, expressed ORNs located in at4 trichoid sensilla on the third antennal segment, is involved in several socio-sexual interactions, including male mating success, mating partner preference, and female aggression to mating pairs (12, 26–29). In males, Or47b senses fatty acid methyl ester and fatty acid that affect mating competition and copulation (30, 31). While Or47b’s role in female aggression is established (12), its involvement in female sexual behaviors is uncertain. For both sexes, the Or47b ORNs project to the VA1v glomeruli, where the VA1v projection neurons receive their signal and project to the mushroom body calyx and lateral horn. Male Or47b neurons link with neurons like aSP5, aSP8, and aSP9, which express a male-specific transcriptional factor FruM (32).

CHC pheromones that function as short-range pheromones are mainly detected by neurons on the forelegs and labellum that express gustatory receptors (GR), ionotropic receptors (IR), or the ppk/DEG-ENaC family of sodium channels (16, 17, 24). CHCs like 7-T and 7,11-HD are sensed by ppk23-expressing M and F cells in the tarsi (33). 7-T and cVA are sensed by M cells expressing ppk23, while 7,11-HD and 7,11-ND are sensed by F cells expressing ppk23, ppk25, and ppk29. In males 7-T or 7,11-HD affects the neuronal activity of the FruM-expressing P1 neurons (34–36). However, how these CHCs signal in the female brain remains unknown.

Sperm ejection is a process through which females can remove the male ejaculate or the mating plug after copulation. This phenomenon was observed in various animal species including feral fowl (37), blacklegged kittiwake (38), and dunnock (39). In the fruit fly, it typically occurs approximately 90 minutes after mating (40). This specific interval, referred to as ‘ejaculate holding period (EHP)’, likely affects sperm usage and fecundity (40, 41). The neurosecretory neurons in the brain pars intercerebralis (PI) that produce diuretic hormone 44 (Dh44), an insect orthologue of the corticotropin-releasing factor, regulate EHP (40). Drosophila females seem to signal the social sexual context through sperm ejection. For instance, when placed in the food patch with the male ejaculate deposited by other females, females exhibited increased likelihoods of egg-laying (42). However, it remains unknown whether the social sexual context influences sperm ejection and EHP.

The pC1 neurons in females, expressing a specific transcriptional factor DsxF, integrate olfactory and auditory cues associated with male courtship (43, 44). The female pC1 neurons in flies, their male counterparts (i.e., P1 neurons), and the ventrolateral subdivision of ventromedial hypothalamus (VMHvl) neurons in mice share conserved circuit configurations and demonstrate functional similarity in coordinating social and sexual behaviors (45, 46). There are Dsx positive 14 pC1 neurons in each brain hemisphere of the brain, responsive to the male sex-pheromone cVA and courtship songs (44, 47).

Connectome analyses identified 10 pC1 neurons, categorized into five subtypes, with pC1a-c associated with mating behaviors and pC1d and e subtypes linked to aggression (47–52). Although direct evidence connecting pC1 neurons to sperm ejection is limited, they are promising candidates for regulating sperm ejection or EHP, because sperm ejection enables females to eliminate the mating plug and male ejaculate, restoring sexual attractiveness (53).

In this study, we demonstrate that two male pheromones, 2-methyltetracosane (2MC) and 7-T, significantly shorten EHP through Or47b neurons and ppk23 neurons, respectively. These pheromonal pathways converge onto pC1 neurons, increasing their cAMP levels. The elevated cAMP in pC1 neurons reduced EHP, similar to the effects of the male pheromones. Elevated cAMP also enhances pC1 neuron excitability, rendering them more responsive to both olfactory and auditory male courtship cues and promoting further mating after earlier removal of the mating plug. Our study introduces a novel behavioral paradigm, shedding light on the intricate molecular and neuronal pathways underlying female sexual plasticity.

Results

Male-induced EHP shortening (MIES) is dependent on olfaction

To investigate the impact of changes in social sexual context on EHP, we conducted a comparison between the EHP of post-mating females isolated from any male presence and those exposed to actively courting, naive wild-type Canton-S (CS) males immediately following copulation (Fig. 1A). Notably, the EHP of females incubated with naive males was approximately 30 minutes shorter than that of females left in isolation after mating (Fig. 1A, 1B). We refer to this phenomenon as ‘male-induced EHP shortening (MIES)’. In contrast, there was little difference in EHP observed between females incubated with virgin females and those isolated after mating (Fig. 1C).

The presence of males shortens the ejaculate holding period (EHP) in females through olfactory or gustatory sensation

A,Schematic of the experimental procedure to measure male-induced EHP shortening (MIES). A female mates with a wild-type Canton-S (CS) male. Immediately after mating, the female is incubated with a naïve male. Typically, females kept alone exhibit an EHP of approximately 90 min, whereas females incubated with a naïve male exhibit an EHP of approximately 60 min. In this study, we refer to this phenomenon as male-induced EHP shortening (MIES).

B-F, Ejaculation holding period (EHP) or ΔEHP of the females of the indicated genotypes, incubated under the indicated conditions after mating. Normalized EHP (ΔEHP) is calculated by subtracting the mean of reference EHP of females kept alone after mating (leftmost column) from the EHP of females incubated with other flies.

Mann-Whitney Test (n.s. p > 0.05; ****p < 0.0001). Gray circles indicate the EHP or ΔEHP of individual females, and the mean ± SEM of data is presented. Numbers below the horizontal bar represent the mean of EHP differences between treatments.

Male fruit flies employ various sensory signals to attract females during courtship (15). To assess the role of the visual signal in MIES, we examined MIES under dim red lighting conditions and observed that limited illumination had a marginal impact on MIES (Fig. 1D). Next, we examined MIES in post-mating females incubated with decapitated CS males. These males could serve as a source for olfactory or gustatory signals, but not for auditory or visual signals. Once again, we observed no reduction in MIES (Fig. 1E). This strongly suggests that olfactory or gustatory cues are the key signals responsible for MIES. This is further supported by the observation that females with a deficiency in the odorant receptor coreceptor (Orco1) exhibited no MIES (Fig. 1F). Thus, it is highly likely that male odorant(s), especially those detected by olfactory receptors (Or), are inducing MIES.

MIES is dependent on Or47b receptor and Or47b-expressing ORNs

In fruit fly antenna, the trichoid sensilla and their associated olfactory receptor neurons are known to detect sex pheromones (54). To explore the contribution of ORNs located in the trichoid sensilla to MIES, we silenced 11 different ORN groups found in the trichoid and intermediate sensilla (55, 56) by expressing either the active or inactive form of Tetanus toxin light chain (TNT) (57). Our findings revealed that silencing ORNs expressing Or13a, Or19a, Or23a, Or47b, Or65c, Or67d or Or88a significantly impacted MIES (Fig. S1A).

We then focused on analyzing Or47b-positive ORNs (Fig. 2A), which, unlike the others, exhibited nearly complete abolition of MIES when silenced. Activation of these neurons with the thermogenetic activator dTRPA1 (58) resulted in a significant EHP shortening, even in the absence of male exposure (Fig. 2B). Subsequently, we addressed whether restoring Orco expression in Or47b ORNs in Orco-deficient females would restore MIES. Our results confirmed that indeed it did (Fig. 2C). To ascertain the necessity of the Or47b receptor gene for MIES, we examined Or47b-deficient females (Or47b2/Or47b3) and observed a complete absence of MIES, whereas heterozygous controls exhibited normal MIES (Fig. 2D). Furthermore, the reintroduction of Or47b expression in Or47b ORNs of Or47b-deficient females nearly completely restored MIES (Fig. 2E). With strength of these observations, we concluded that MIES hinges on the Or47b receptor gene and Or47b-expressing ORNs.

Or47b and Or47b-positive ORNs are required for MIES

A, C-E, ΔEHP of females of the indicated genotypes, incubated with or without naive males. Female genotypes are as follows from left to right: (A) control (Or47b>TNTinactive), Or47b ORN silencing (Or47b>TNTactive); (C) Orco mutant (Orco1/Orco1), Orco rescue in Or47b ORNs of Orco mutant (Orco1/Orco1; Or47b>Orco); (D) control 1 (Or47b2/+), control 2 (Or47b3/+), Or47b mutant (Or47b2/Or47b3); (E) Or47b mutant (Or47b2/Or47b2), Or47b rescue (Or47b>Or47b; Or47b2/Or47b2).

B, Thermal activation of Or47b-positive ORNs shortens EHP in the absence of naive males. ΔEHP is calculated by subtracting the mean of reference EHP of females incubated at 21°C control conditions from the EHP of individual females. Female genotypes are as follows from left to right: control 1 (Or47b-Gal4/+), control 2 (UAS-dTRPA1/+), Or47b>dTRPA1 (Or47b-Gal4/UAS-dTRPA1). Mann-Whitney Test (n.s. p > 0.05; *p <0.05; **p <0.01; ****p < 0.0001). Gray circles indicate the ΔEHP of individual females, and the mean ± SEM of data is presented. Gray circles with dashed borders indicate ΔEHP values beyond the axis limits (>90 or <-90 min). Numbers below the horizontal bar represent the mean of EHP differences between treatments.

2-Methyltetracosane (2MC) induces MIES through Or47b and Or47b ORNs

Previous studies have shown that methyl laurate (ML) and palmitoleic acid (PA) can activate Or47b ORNs only in the presence of a functional Or47b gene (30, 31). However, in our investigation, none of these odorant ligands for the Or47b receptor induced significant EHP shortening, even up to 1440 ng (Fig. S2). This prompted us to seek a new pheromone capable of activating Or47b ORNs and therefore shortening EHP.

Oenocytes are known to produce a significant portion, albeit not all, of cuticular hydrocarbons or pheromones. To ascertain whether the male pheromone responsible for MIES is produced by oenocytes, we conducted experiments to determine whether incubation with females possessing masculinized oenocytes would lead to EHP shortening (Fig. S3A). Indeed, our findings confirmed that females with male oenocytes significantly shortens EHP, strongly suggesting that male oenocytes serve as a source for the MIES pheromone. Unexpectedly, however, incubation with males possessing feminized oenocytes also resulted in significant EHP shortening (Fig. S3A). This raises the possibility that oenocytes may not be the sole source of the MIES pheromone, implying the involvement of more than one pheromone, for instance one from oenocytes and another from an alternate source, in MIES.

The genus Drosophila exhibits distinct CHC profiles, with certain CHC components shared among closely related species (59). We found that incubation with males from other closely related species, such as D. similans, D. sechellia, and D. erecta, also induced EHP shortening, whereas incubation with D. yakuba males did not (Fig. S3B). On a search for a male-specific pheromone present in D. melanogaster, D. similans, D. sechellia, and D. erecta, but not in D. yakuba (30, 60), we identified 2MC, which shortens EHP within a physiologically relevant and narrow concentration range (Fig. 3A; Fig. S4). Moreover, EHP shortening induced by 2MC was not observed in females deficient in Orco or Or47b (Fig. 3B, 3C), but it was restored when Orco expression is reinstated in Or47b ORNs in Orco-deficient mutants (Fig. 3D). 2MC was found mainly in males, but not in virgin females (30). Our behavioral observations strongly indicate that 2MC acts as an odor ligand for Or47b and shortens EHP through this receptor.

2-methyltetracosane (2MC) can induce EHP shortening through Or47b A-D,

Δ EHP of mated females of the indicated genotypes, incubated in solvent vehicle or 2MC. Mated females were incubated with a piece of filter paper perfumed with vehicle (-) or 750 ng 2MC (+). Female genotypes are as follow: w1118, (B) Orco mutant (Orco1/Orco1), (C) Or47b mutant (Or47b2/Or47b2), (D) Gal4 control (Or47b-Gal4/+; Orco1/Orco1), UAS control (UAS-Orco/+; Orco1/Orco1), Orco rescue in Or47b ORN (Orco1/Orco1; Or47b-Gal4/UAS-Orco).

A-C, Mann-Whitney Test (n.s. p > 0.05; *p <0.05), D, One-way analysis of variance (ANOVA) test (n.s. p > 0.05; *p < 0.05). Gray circles indicate the ΔEHP of individual females and the mean ± SEM of data is presented. Normalized EHP (ΔEHP) is calculated by subtracting the mean of reference EHP of females incubated with vehicle (A-C) or mean of Gal4 control and UAS control female incubated with vehicle (D) after mating from the EHP of females incubated with chemical perfumed paper. Gray circles with dashed borders indicate ΔEHP values beyond the axis limits (>90 or <-90 min). Numbers below the horizontal bar represent the mean of EHP differences between treatments.

7-Tricosene (7-T) shortens EHP through ppk23 neurons

In contrast to incubation with virgin females, incubation with mated females resulted in a significant shortening in EHP (Fig. 4A). Unlike virgin females, mated females carry male pheromones, including 7-T and cVA, transferred during copulation (53). This raised the possibility that that these male pheromones might also induce EHP shortening. Indeed, our experiments revealed that incubation with a piece of filter paper infused with 150 or 375 ng of 7-T significantly shortened EHP. Conversely, incubation with cVA and 7-pentacosene, a related CHC, did not produce the same effect (Fig. 4B, 4C; Fig. S5A-5B). The concentrations of 7-T capable of inducing EHP shortening appear to be physiologically relevant. 7-T was found in quantities of 432 ng in males (61), 25 ng in virgin females, and 150 ng in mated females (53).

7-Tricosene present in mated females or naive males shorten EHP via ppk23 neurons A-D,

ΔEHP of females of the indicated genotypes, incubated with mated females (A), a piece of filter paper perfumed with 150 ng 7-T (B) or 200 ng cVA (C), or naive males (D) immediately after mating. Female genotypes are: (A-C) w1118, (D) control (ppk23-Gal4/UAS-TNTinactive), ppk23 silencing (ppk23-Gal4/UAS-TNTactive).

Unpaired t-Test (n.s. p > 0.05; * p < 0.05). Gray circles indicate the ΔEHP of individual females, and the mean ± SEM of data is presented. Gray circles with dashed borders indicate ΔEHP values beyond the axis limits (>90 or <-90 min). Numbers below the horizontal bar represent the mean of EHP differences between treatments.

Although the receptors for 7-T remain unknown, ppk23-expressing tarsal neurons have been shown to sense these compounds and regulate sexual behaviors in males and females (23, 62, 63). Subsequently, we silenced ppk23 neurons, and as a result, MIES is almost completely abolished, underscoring the pivotal role of 7-T in MIES (Fig. 4D). However, DEG/ENac channel genes expressed in ppk23 neurons, including ppk23 and ppk29, were found to be dispensable for MIES (Fig. S5C-5E). This aligns with previous observations that neither ppk23 deficiency nor ppk28 deficiency replicates the sexual behavioral defects caused by silencing ppk23 neurons (64).

A pC1 neuron subset regulates EHP and MIES

The neuropeptide Dh44 determines timing of sperm ejection or EHP (40). The same study found that Dh44 receptor neurons involved in EHP regulation also express a sexually dimorphic transcriptional factor gene doublesex (dsx). A recent study has revealed that pC1 neurons, a specific subgroup of dsx-expressing central neurons in the female brain, indeed express Dh44 receptors (47). With these findings, we set out to investigate the roles of pC1 neurons in the regulation of EHP and MIES. The pC1 neurons comprise five distinct subtypes. Among these, the pC1a, b, and c subtypes have been associated with mating receptivity (47, 48), while the remaining pC1d and e subtypes have been linked to female aggression (48, 52). To investigate the impact of these subtypes on EHP, we employed GtACR1, an anion channel activated by blue light in the presence of all-trans-retinal (ATR), to silenced specific pC1 subtypes immediately after mating. Our experiments revealed that silencing the pC1 subset comprising pC1a, b and c subtype with GtACR1 led to an increase in EHP (Fig. 5A), whereas silencing the pC1d and e subtypes had a limited effect on EHP (Fig. 5B). Furthermore, we dissected the roles of pC1b and c along with pC1a separately.

pC1 neuron subset comprising pC1b and c subtypes regulates EHP in response to 2MC and 7-T, both of which upregulate cAMP activity and excitability of pC1 neurons

A-D, Optogenetic silencing of a pC1 neuron subset comprising pC1b and c subtypes increases EHP. Females of the indicated genotypes were cultured on food with or without all-trans-retinal (ATR). ΔEHP is calculated by subtracting the mean of reference EHP of females cultured in control ATR-food from the EHP of individual test females. Female genotypes are: (A) pC1a,b,c>GtACR1 (pC1-S-Gal4/UAS-GtACR1), (B) pC1d,e>GtACR1 (pC1-A-Gal4/UAS-GtACR1), (C) pC1a>GtACR1 (pC1a-split-Gal4 /UAS-GtACR1), and (D) pC1b,c>GtACR1 (Dh44-pC1-Gal4/UAS-GtACR1). Gray circles indicate the ΔEHP of individual females, and the mean ± SEM of data is presented. Gray circles with dashed borders indicate ΔEHP values beyond the axis limits (>120 min). Mann-Whitney Test (n.s. p > 0.05; *p <0.05; ****p < 0.0001). Numbers below the horizontal bar represent the mean of EHP differences between treatments.

E, Relative CRE-Luciferase reporter activity of pC1 neurons in mated females of the indicated genotypes, incubated with a piece of filter paper perfumed with solvent vehicle control or the indicated odorants. To calculate the relative luciferase activity, we set the average luminescence unit values of female incubated with the vehicle to 100%. One-way ANOVA test (n.s. p > 0.05; ***p < 0.001; ****p < 0.0001). Gray circles indicate the relative luciferase activity (%) of individual females, and the mean ± SEM of data is presented.

F, Optogenetic production of cAMP in the pC1 b and c neurons shortens EHP, whereas the same treatment in pC1a or pC1d and e neurons does not. ΔEHP is calculated by subtracting the mean of reference EHP of females incubated in the control illumination (Dim light), which does not activate a photoactivatable adenylate cyclase (PhotoAC), from the EHP of individual test females. Mann-Whitney Test (n.s. p > 0.05, ****p < 0.0001).

G, Optogenetic production of cAMP increases excitability of pC1 neurons transiently. Left, schematic of the experimental procedure. Right, peak ΔF/F in the LPC projections of pC1 neurons from freshly mated females in response to the pheromone cVA, before and after photoactivation of PhotoAC expressed in pC1 neurons. Calcium response was measured at specific time points: after 1 minute (Blue dots and box, 1 TAA = 1 Time After Activation) or 10 minutes (Purple dots and box, 10 TAA) after activation. Repeated measures (RM) one-way ANOVA test with the Geisser-Greenhouse correction followed by Tukey’s multiple comparisons test (*p < 0.05; ***p < 0.001; ****p < 0.0001).

H, Left, schematic of the experimental procedure. Right, re-mating rate of females during optogenetic cAMP production in pC1b and c, scored as the percentage of females that copulate with naive male within 6 h after end of first mating. Female genotypes are control (+/UAS-PhotoAC), pC1b,c>UAS-PhotoAC (Dh44-pC1-Gal4/UAS-PhotoAC). Chi-square test (*p <0.05).

We generated a subtype-specific split-Gal4 for pC1a and found that, as expected, silencing pC1a using this split-Gal4 nearly completely suppressed mating receptivity (Fig. S6). However, the silencing pC1a alone did not result in increased EHP, suggesting a marginal role of the pC1a subtype in EHP regulation (Fig. 5C). In contrast, concomitant silencing of both pC1b and pC1c significantly increased EHP by 56 + 6.9 minutes (Fig. 5D). Currently, we lack the genetic tools required to further distinguish the roles of pC1b and pC1c subtypes individually.

Our recent research has uncovered that pC1 neurons exhibit elevated cAMP activity during sexual maturation, with this increase in cAMP being closely linked to heightened excitability of the pC1 neurons (47). The same study also revealed that a mating signal (i.e., sex peptide in the male seminal fluid) reduces cAMP activity in pC1 neurons. Thus, we hypothesized that male odorants responsible for inducing MIES, such as 2MC or 7-T, would elevate cAMP activity in pC1b and pC1c neurons in freshly mated females. This, in turn, would lead to increased excitability of pC1 neurons and, as a consequence, a reduction in the EHP. To monitor the cAMP activity in these neurons, we prepared females that selectively produce the CRE-Luciferase reporter gene in pC1b and pC1c neurons. Indeed, when exposed to 2MC or 7-T, pC1b and pC1c neurons exhibited a significant increase in CRE-luciferase activity, indicating that these neurons produce higher levels of cAMP in response to these odorants (Fig. 5E). Notably, the CRE-Luciferase activity appeared to reach its peak at specific odorant concentrations that induced a significant shortening of EHP (Fig. S7).

In contrast, when examining other pC1 subsets, such as pC1a, and pC1d and e, we detected no sign of increased CRE-Luciferase reporter activity upon exposure to 2MC or 7-T treatment (Fig. 5E). It is worth noting that CRE-luciferase reporter activity in the pC1a neurons appears to be dependent on the mating status, as it reaches levels similar to those of pC1b and pC1c neurons in virgin females (Fig. S8). This observation aligns well with connectome data, which indicates that SAG neurons, responsible for relaying SP-dependent mating signals, primarily establishes synapses into pC1a subtype and to a much lesser extent into other pC1 subtypes (49).

Having shown that MIES-inducing male odorants, 2MC or 7-T, increase cAMP activity in pC1b and pC1c neurons from mated females, we next asked whether this induced elevation of cAMP levels in pC1b and pC1c would shorten EHP, leading to MIES. We employed the photoactivatable adenylate cyclase (PhotoAC), which augments cellular cAMP levels upon exposure to light. Indeed, the induced elevation of cAMP activity in pC1b and pC1c significantly shortened EHP, whereas the same treatment applied to pC1a or pC1d and pC1e had no such effect (Fig. 5F). This further underscores the pivotal role of pC1b and pC1c in EHP regulation.

Next, we asked whether the expression of Dh44R1 and Dh44R2, GPCRs that increase cellular cAMP in response to their ligand Dh44, in pC1b and pC1c neurons is necessary for MIES. However, double knockdown of Dh44R1 and Dh44R2 in pC1 neurons seemed to have a limited impact on MIES (Fig. S9). This suggests that Dh44R signaling in pC1 neurons is not essential for the regulation of EHP or MIES, opening up the possibility that other GPCRs may be involved in up-regulating cAMP levels in pC1 neurons in response to 2MC or 7-T.

Lastly, we investigated how the increased cAMP activity affects physiological activity of pC1 neurons. pC1 neurons from virgin females exhibit robust Ca2+ transients in response to male courtship cues, such as the male pheromone cVA and courtship pulse song (44). In contrast, those from mated females display significantly diminished Ca2+ transients (47). Shortly after mating, a decrease in pC1 responsiveness to cVA was observed. However, immediately following the activation of PhotoAC in pC1 neurons, pC1 neurons in freshly mated females became more excitable and exhibited stronger Ca2+ transients in response to cVA (Fig. 5G). It is important to note that this PhotoAC-induced increase in pC1 excitability is transient and diminishes rapidly within 10 minutes (Fig. 5G). Together, these findings suggest that the increased cAMP levels in pC1 neurons would not only promote MIES but also facilitate re-mating in post-mating females, which typically engage in additional mating at a low frequency. To test this hypothesis, we examined the re-mating frequency of freshly mated females paired with naive males while inducing a cAMP increase in pC1 neurons. As expected, PhotoAC activation in pC1b and c neurons substantially increased the re-mating rate compared to the control group (Fig. 5H). Therefore, we concluded that male odorants, which stimulate cAMP elevation in pC1 neurons. expedite the removal of mating plug, consequently leading to increased instances of re-mating.

Discussion

Males employ a diverse range of strategies to enhance their reproductive fitness. One such strategy involves the formation of a ‘mating plug’, a mechanism that prevents females from engaging in further mating and consequently increases fertilization success rates (65–67). As a means of intra-sexual competition, rival males often promote the removal or precocious expulsion of the mating plug. This behavior is driven by the fact that polyandrous females often eliminate the mating plug to engage in additional mating with males possessing superior traits or higher social status than their previous partners (37, 68). In the dunnock (Prunella modularis), a small European passerine bird, the male often engages in cloacal pecking of copulated females, inducing the expulsion of prior mate’s sperm and mating plug, thereby increasing their chance of successful mating (39). In this study, we discovered that in D. melanogaster, when kept with actively courting males, freshly mated females exhibit an earlier removal of the mating plugs or a shorter EHP. This behavior is primarily induced by the stimulation of females via male sex pheromones. Furthermore, our study has uncovered the conserved neural circuitry that processes male courtship cues and governs mating decisions play an important role in regulating this behavior. By delving into the molecular and neuronal mechanisms underlying MIES, our study provides valuable insights into the broader aspect of behaviors induced by social sexual context changes.

Our findings highlight the involvement of the Or47b receptor and Or47b ORNs in MIES. These OR and ORNs have been linked to a range of social and sexual behaviors in both males and female fruit flies (12, 26–31). Methyl laurate and trans-palmitoleic acid are odor ligands for Or47b, which explain many of these functions particularly in males (30, 31). In this study, we provide compelling evidence that 2MC induces EHP shortening via both Or47b receptor and Or47b ORNs, suggesting that 2MC functions as an odor ligand for Or47b. Notably, a gas chromatography–mass spectrometry (GC-MS) analysis of cuticular hydrocarbons of 4-day-old wild type D. melanogaster indicated the presence of 2MC solely in males, excluding females (30). Surprisingly, however, unlike 2MC, neither methyl laurate nor trans-palmitoleic acid influences EHP. The reason of this paradoxical results remains unclear. Plausible interpretation is that the EHP shortening induced by 2MC may require not only Or47b but also other as-yet-unidentified ORs.

With the establishment of a behavioral and cellular assessment for 2MC activity, the search for additional odorant receptors responsive to 2MC is now feasible. An other important avenue for further research is whether 2MC can also elicit behaviors that were previously associated with methyl laurate or transpalmitoleic acid, such as the promotion of male copulation and courtship (30, 31).

We observed that both 2MC and 7-T exhibit both cellular and behavioral activity within a specific concentration range (Fig. S4, S5, S7). This observation is of particularly interest, given the multitude of environmental and biological factors that influence the levels of 2MC and 7-T, potentially affecting the capacity of males to induce MIES. For instance, exposure to low temperatures during development has been linked to increased production of both 2MC and 7-T (69). Similarly, the mutation of the desiccation stress gene CG9186, which encodes a protein associated with lipid droplet, has been found to impact 2MC levels (70). Furthermore, 2MC levels rise with the aging of males (71). Thus, we propose that the levels of 2MC and possibly 7-T may serve as indicators of male’s age and their resilience against environmental stresses in a complex and non-linear manner.

In mated females, treatment with 2MC or 7-T increases cAMP activity in pC1b,c neurons but not in pC1a neurons. In contrast, pC1a neurons in virgin females are fully responsive to both male pheromones, showing a cAMP activity that is similar to that of pC1b,c neurons (Fig. S8). The absence of cAMP activity in pC1a neurons in mated females likely results from the mating signal (i.e., sex peptide) silencing pC1a neurons. Connectome and electrophysiology data support this interpretation, as SAG neurons, which relay sex peptide signals, exhibit the strongest synaptic connection with pC1a subtype among five pC1 subtypes (49). However, SAG neuron activity may also influence pC1c neurons, as they also have substantial synaptic connections with pC1c neurons in hemibrain (72). Notably, induced activation of SAG neurons significantly shortens EHP (Fig. S10), suggesting they regulate EHP through pC1c neurons.

We found that increased cAMP levels causes pC1b,c neurons in mated females, which typically do not respond to male courtship cues like cVA and pulse song, to become responsive, showing strong Ca2+ transients. Since pC1b,c neurons play a role in generating sexual drive and increasing female receptivity to male courtship, the 2MC or 7-T-induced increases in cAMP likely govern the removement of the mating plug and engagement in further mating of mated females. This observation well aligns with the previous report that mating reduces sensitivity of Or47b ORNs, which we find responsive to 2MC, leading to the increased preference for pheromone-rich males after mating (29). Moreover, the finding that 2MC and 7-T induce cAMP activity in pC1b,c neurons in virgin females suggests that virgin females may also use 2MC and 7-T as odorant cues to assess male quality during their first mating. Indeed, females seem to evaluate male quality with the amount of 7-T, as increased 7-T promotes mating receptivity and shortens mating latency (20).

Physiological factors like the nutritional status of females before mating and the nutritional status of their mating partners have been shown to influence EHP (73), and therefore potentially MIES. Hence, it is highly probable that MIES is regulated by additional central neurons such as Dh44-PI neurons that regulate these processes (40). However, it remains unclear whether and how Dh44-PI neurons and pC1 neurons interact to modulate EHP and MIES. The observation that double knockdown of Dh44R1 and Dh44R2 has only a marginal effect on MIES suggests that Dh44-PI neurons may also function independently of pC1 neurons, suggesting the possibility that multiple independent central circuits may contribute to the production of MIES.

Our initial screening of ORNs responsible for MIES revealed the involvement of Or47b ORNs, as well as several other ORNs. In addition to 2MC, which operates through Or47b expressing ORNs, our findings indicate that 7-T and ppk23 neurons capable of detecting 7-T also play a role in MIES induction. In D. melanogaster and other related species, food odors typically serve as long-range signals that attract both males and females (74, 75), implying that certain food odors might also influence EHP (42). The involvement of multiple odorants and ORNs in EHP regulation implicates that pC1 neurons can process various cues, not restricted to those associated with mating behaviors, including food odors. Future studies will explore the full spectrum of odorants processed by pC1 neurons in EHP regulation.

In conclusion we have identified a circuit that, via the detection of a novel male pheromone, potentially signals male quality and governs the female decision to remove the mating plug of its last mate and mate again.

Materials and Methods

Fly care

Flies were cultured on a standard medium composed of dextrose, cornmeal, and yeast, under room temperature in a 12hr : 12hr light:dark cycle (40, 73). Behavioral assays were performed at 25 °C, except for thermal activation experiment with dTRPA1. Virgin males and females were collected immediately after eclosion. Males were individually aged for 4-6 days, while females were aged in groups of 15–20. For EHP and mating assays, females were aged for 3-4 days. Assays were performed at Zeitgeber time (ZT) 3:00– 11:00, and were repeated at least on three separate days.

Fly stocks

The following stocks are from the Bloomington Drosophila Stock Center (BDSC), the Vienna Drosophila Resource Center (VDRC): Canton S (CS) (RRID: BDSC_64349), w1118 (VDRC #60000), R71G01 (pC1-Gal4) (RRID: BDSC_39599), Orco1 (RRID: BDSC_23129), Or13a-Gal4 (RRID: BDSC_9946), Or19a-Gal4 (RRID: BDSC_9948), Or23a-Gal4 (RRID: BDSC_9955), Or43a-Gal4 (RRID: BDSC_9974), Or47b-Gal4 (RRID: BDSC_9983), Or47b-Gal4 (RRID: BDSC_9984), Or65a-Gal4 (RRID: BDSC_9993), Or65b-Gal4 (RRID: BDSC_23901), Or65c-Gal4 (RRID: BDSC_23903), Or67d-Gal4 (RRID: BDSC_9998), Or83c-Gal4 (RRID: BDSC_23131), Or88a-Gal4 (RRID: BDSC_23137), UAS-Or47b (RRID: BDSC_76045), Or47b2/2 (RRID: BDSC_51306), Or47b3/3 (RRID: BDSC_51307), UAS-TNT active (RRID: BDSC_28837), UAS-TNT inactive (RRID: BDSC_28839), UAS-dTRPA1 (RRID: BDSC_26263), UAS-CsChrimson (RRID: BDSC_55135), UAS-GCaMP6m (RRID: BDSC_42748), R52G04-AD (RRID: BDSC_71085), SAG-Gal4 (VT50405) (RRID:Flybase_FBst0489354, VDRC #200652), UAS-Dh44R1-RNAi (RRID:Flybase_FBst0482273, VDRC #110708), UAS-Dh44R2-RNAi (RRID:Flybase_FBst0465025, VDRC #43314), UAS-Dicer2 (VDRC #60007). The following stocks are reported previously: PromE(800)-Gal4 (59), UAS-FLP, CRE-F-luc (76), LexAop-FLP (77), UAS-CsChrimson (78), UAS-GtACR1 (79), UAS-PhotoAC (PACα) (80), pC1-A (48), pC1-S (48), Dh44-pC1-Gal4 (47), ppk23-Gal4, ppk23-, ppk28-, ppk29-(62), and Orco-Gal4, UAS-EGFP-Orco (81). pC1a-split-Gal4 is generated by combining R52G04-AD (RRID: BDSC_71085) and dsx-DBD (49). Drosophila subgroups species are obtained from the EHIME-Fly Drosophila stock center. To enhance knock-down efficiency, RNAi experiments were performed using flies carrying UAS-Dicer2 (VDRC #60007).

Chemical information

All trans-retinal (Cat# R2500), methyl laurate (Cat# W271500), and Triton™ X-100 (Cat# X100) were obtained from Sigma-Aldrich (St. Louis, MO, USA). The following chemicals are from the Cayman chemical (Ann Arbor, MI, USA): 7(Z)-Tricosene (CAS No. 52078-42-9, Cat# 9000313), 7(Z)-Pentacosene (CAS No. 63623-49-4, Cat# 9000530), trans-palmitoleic acid (CAS No. 10030-73-6, Cat# 9001798), 11-cis-vaccenyl acetate (cVA) dissolved in EtOH (CAS No. 6186-98-7, Cat# 10010101). 2-methyltetracosane (>98%, purity) was custom-synthesized by KIP (Daejeon, Korea).

Behavior assays

For mating behavior assays, we followed the procedures described previously (82). Individual virgin females and naive CS males were paired in chambers with a diameter of 10 mm and were recorded using a digital camcorder (SONY, HDR-CX405 or Xiaomi, Redmi Note 10) for either 30 min or 1 hour for the mating assay and 6 hours for the re-mating assay. In the re-mating assay, females that completed their initial mating within 30 min were subsequently paired with naive CS males.

To measure EHP, which is defined as the time elapsed between the end of copulation and sperm ejection, we used the following procedure: Virgin females were paired individually with CS males in 10-mm diameter chambers. Following copulation, females were transferred to new chambers, either with or without a CS male or a piece of filter paper treated with pheromones, and their behavior was recorded using a digital camcorder (SONY, HDR-CX405). Typically, females that completed copulation within 30 minutes were used for analysis. To present the pheromone, females were individually kept in 10-mm diameter chambers containing a piece of Whatman filter paper (2 mm x 2 mm) that had been treated with 0.5 μl of the pheromone solution and left to air dry for 1 minute. For thermal activation experiments, females were incubated at the specified temperatures immediately after copulation ended. For light-activation experiments, a custom-made light activation setup utilizing a ring of 104 multi-channel LED lights (NeoPixel, Cat# WS2812; Red light, 620-625 nm, 390-420 mcd; Green light, 522-525 nm, 660-720 mcd; Blue light, 465-467 nm, 180-200 mcd) was employed. Females were individually placed in 10-mm diameter chambers, and the chamber was illuminated with light at an intensity of 1100 lux across the chamber, as measured by a lightmeter HS1010, during the assay. Flies used in these experiments were prepared by culturing them in food containing vehicle (EtOH) or 1 mM all-trans-retinal (ATR) immediately after eclosion. They were kept in complete darkness for 3-4 days until the assay was conducted.

Calcium imaging

We followed the procedures described previously (47, 83). Following copulation, freshly mated female fly was temporally immobilized using ice anesthesia, and its head was attached to a custom made thin metal plate with a 1mm diameter hole using photo-curable UV glue (ThreeBond, A16A01). An opening in the fly’s head was generated using a syringe needle under saline (108 mM NaCl, 5 mM KCl, 2 mM CaCl2, 8.2 mM MgCl2, 4 mM NaHCO3, 1 mM NaH2PO4, 5 mM trehalose, 10 mM sucrose, 5 mM HEPES pH 7.5). Imaging was performed using Zeiss Axio Examiner A1 microscope equipped with an electron multiplying CCD camera (Andor Technology, LucaEM R 604M) and a LED light-source (CoolLED, Precis Excite). Imaging analysis was used Metamorph software (Molecular Devices, RRID:SCR_002368). To deliver the male pheromone using an airflow, we used the Syntech Stimulus Controller (Type CS-55). 2 μl of pheromone solution was applied to a piece of Whatman filter paper (2 mm x 1 mm), which was then inserted into a glass Pasteur pipette after solvent evaporation.

Luciferase assay

We followed the procedures described previously (47, 76). 3-day-old virgin females or freshly mated females were used for assay. A group of three fly heads, kept at -80°C, was homogenized using cold homogenization buffer (15 mM HEPES, 10 mM KCl, 5 mM MgCl2, 0.1 mM EDTA, 0.5 mM EGTA).

Luciferase activity was measured by using beetle Luciferin, potassium Salt (Promega, Cat# E1603) and a microplate Luminometer (Berthold technologies, Centro XS3 LB 960), following manufacturer’s instructions. To present the odorant, flies were placed in 10-mm diameter chambers containing a piece of Whatman filter paper (4 mm x 6 mm) that had been treated with 1 μl of the pheromone solution and left to air dry for 1 minute.

Immunohistochemistry

3–5-day-old virgin female flies were dissected in phosphate-buffered saline (PBS) and fixed for 30 minutes at room temperature in 4% paraformaldehyde in PBS. Following fixation, the brains were thoroughly washed in PBST (0.1% Triton™ X-100 in PBS) and then blocked using 5% normal goat serum in PBST. After blocking, the brains were incubated with the primary antibody in PBST for 48 hours at 4 °C, washed by PBST and then incubated with the secondary antibody in PBST for 24 hours at 4 °C. The samples were washed three times with PBST and once with PBS before being mounted in Vectashield (Vector Laboratories, Cat# H-1000). Antibodies used were rabbit anti-GFP (1:1000; Thermo Fisher Scientific, Cat# A-11122, RRID:AB_221569), mouse anti-nc82 (1:50; Developmental Studies Hybridoma Bank, Cat# Nc82; RRID: AB_2314866), Alexa 488-conjugated goat anti-rabbit (1:1000; Thermo Fisher Scientific, Cat# A-11008, RRID:AB_143165), Alexa 568-conjugated goat anti-mouse (1:1000; Thermo Fisher Scientific, Cat# A-11004, RRID:AB_2534072). Brain images were acquired using Zeiss LSM 700/Axiovert 200M (Zeiss) and processed using Fiji (https://imagej.net/software/fiji/downloads, RRID:SCR_002285)

Color-depth MIP based anatomy analysis

A stack of confocal images from pC1a-split-Gal4>UAS-myr-EGFP adult female brains stained with anti-GFP and anti-nc82 were used. Images were registered onto the JRC2018 unisex brain template (84). using the Computational Morphometry Toolkit (CMTK, https://github.com/jefferis/fiji-cmtk-gui). Color depth MIP masks of pC1a-split-Gal4 neurons and pC1a (ID, 5813046951) in Hemibrain (72) (Fig. S6A) were generated using the ColorMIP_Mask_Search plugin (85) for Fiji (https://github.com/JaneliaSciComp/ColorMIP_Mask_Search) and NeuronBridge (86) (https://neuronbridge.janelia.org/). Similarity score and rank were calculated using NeuronBridge.

Statistical analysis

Statistical analysis was conducted using GraphPad Prism 9 (Graphpad, RRID:SCR_002798), with specific details on each statistical method provided in the figure legends.

Supporting information

eLife Assessment

This fundamental work unravels how female Drosophila can assess their social context via chemosensory cues and modulate the sperm storage process after copulation accordingly. A convincing set of rigorous experiments uncovers specific pheromones that influence the excitability of the female brain receptivity circuit and their propensity to discard inseminate from a mating. This insight into neuronal mechanisms of sexual behavior plasticity is of general interest to scientists working in the fields of animal behavior, neuroscience, evolution, and sexual selection, as well as insect chemosensation and reproduction.

Reviewer #1 (Public Review):

Yun et al. examined the molecular and neuronal underpinnings of changes in Drosophila female reproductive behaviors in response to social cues. Specifically, the authors measure the ejaculate-holding period, which is the amount of time females retain male ejaculate after mating (typically 90 min in flies). They find that female fruit flies, Drosophila melanogaster, display shorter holding periods in the presence of a native male or male-associated cues, including 2-Methyltetracosane (2MC) and 7-Tricosene (7-T). They further show that 2MC functions through Or47b olfactory receptor neurons (ORNs) and the Or47b channel, while 7-T functions through ppk23 expressing neurons. Interestingly, their data also indicates that two other olfactory ligands for Or47b (methyl laurate and palmitoleic acid) do not have the same effects on the ejaculate-holding period. By performing a series of behavioral and imaging experiments, the authors reveal that an increase in cAMP activity in pC1 neurons is required for this shortening of the ejaculate-holding period and may be involved in the likelihood of remating. This work lays the foundation for future studies on sexual plasticity in female Drosophila.

The conclusions of this paper are mostly supported by the data, but aspects of the lines used for individual pC1 subtypes and visual contributions as well as the statistical analysis need to be clarified.

(1) The pC1 subtypes (a - e) are delineated based on their morphology and connectivity. While the morphology of these neurons is distinct, they do share a resemblance that can be difficult to discern depending on the imaging performed. Additionally, genetic lines attempting to label individual neurons can easily be contaminated by low-level expression in off-target neurons in the brain or ventral nerve cord (VNC), which could contribute to behavioral changes following optogenetic manipulations. In Figures 5C - D, the authors generated and used new lines for labeling pC1a and pC1b+c. The line for pC1b+c was imaged as part of another recent study (https://doi.org/10.1073/pnas.2310841121). However, similar additional images of the pC1a line (i.e. 40x magnification and VNC expression) would be helpful in order to validate its specificity.

(2) The author's experiments examining olfactory and gustatory contributions to the holding period were well controlled and described. However, the experiments in Figure 1D examining visual contributions were not sufficiently convincing as the line used (w1118) has previously been shown to be visually impaired (Wehner et al., 1969; Kalmus 1948). Using another wild-type line would have improved the authors' claims.

(3) When comparisons between more than 2 groups are shown as in Figures 1E, 3D, and 5E, the comparisons being made were not clear. Adding in the results of a nonparametric multiple comparisons test would help for the interpretation of these results.

Reviewer #2 (Public Review):

The work by Yun et al. explores an important question related to post-copulatory sexual selection and sperm competition: Can females actively influence the outcome of insemination by a particular male by modulating the storage and ejection of transferred sperm in response to contextual sensory stimuli? The present work is exemplary for how the Drosophila model can give detailed insight into the basic mechanism of sexual plasticity, addressing the underlying neuronal circuits on a genetic, molecular, and cellular level.

Using the Drosophila model, the authors show that the presence of other males or mated females after mating shortens the ejaculate-holding period (EHP) of a female, i.e. the time she takes until she ejects the mating plug and unstored sperm. Through a series of thorough and systematic experiments involving the manipulation of olfactory and chemo-gustatory neurons and genes in combination with exposure to defined pheromones, they uncover two pheromones and their sensory cells for this behavior. Exposure to the male-specific pheromone 2MC shortens EHP via female Or47b olfactory neurons, and the contact pheromone 7-T, present in males and on mated females, does so via ppk23 expressing gustatory foreleg neurons. Both compounds increase cAMP levels in a specific subset of central brain receptivity circuit neurons, the pC1b,c neurons. By employing an optogenetically controlled adenyl cyclase, the authors show that increased cAMP levels in pC1b and c neurons increase their excitability upon male pheromone exposure, decrease female EHP, and increase the remating rate. This provides convincing evidence for the role of pC1b,c neurons in integrating information about the social environment and mediating not only virgin but also mated female post-copulatory mate choice.

Understanding context and state-dependent sexual behavior is of fundamental interest. Mate behavior is highly context-dependent. In animals subjected to sperm competition, the complexities of optimal mate choice have attracted a long history of sophisticated modelling in the framework of game theory. These models are in stark contrast to how little we understand so far about the biological and neurophysiological mechanisms of how females implement post-copulatory or so-called "cryptic" mate choice and bias sperm usage when mating multiple times.

The strength of the paper is decrypting "cryptic" mate choice, i.e. the clear identification of physiological mechanisms and proximal causes for female post-copulatory mate choice. The discovery of peripheral chemosensory nodes and neurophysiological mechanisms in central circuit nodes will provide a fruitful starting point to fully map the circuits for female receptivity and mate choice during the whole gamut of female life history.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public Review):

Yun et al. examined the molecular and neuronal underpinnings of changes in Drosophila female reproductive behaviors in response to social cues. Specifically, the authors measure the ejaculate-holding period, which is the amount of time females retain male ejaculate after mating (typically 90 min in flies). They find that female fruit flies, Drosophila melanogaster, display shorter holding periods in the presence of a native male or male-associated cues, including 2-Methyltetracosane (2MC) and 7-Tricosene (7-T). They further show that 2MC functions through Or47b olfactory receptor neurons (ORNs) and the Or47b channel, while 7-T functions through ppk23 expressing neurons. Interestingly, their data also indicates that two other olfactory ligands for Or47b (methyl laurate and palmitoleic acid) do not have the same effects on the ejaculate-holding period. By performing a series of behavioral and imaging experiments, the authors reveal that an increase in cAMP activity in pC1 neurons is required for this shortening of the ejaculate-holding period and may be involved in the likelihood of remating. This work lays the foundation for future studies on sexual plasticity in female Drosophila.

The conclusions of this paper are mostly supported by the data, but aspects of the lines used for individual pC1 subtypes and visual contributions as well as the statistical analysis need to be clarified.

(1) The pC1 subtypes (a - e) are delineated based on their morphology and connectivity. While the morphology of these neurons is distinct, they do share a resemblance that can be difficult to discern depending on the imaging performed. Additionally, genetic lines attempting to label individual neurons can easily be contaminated by low-level expression in off-target neurons in the brain or ventral nerve cord (VNC), which could contribute to behavioral changes following optogenetic manipulations. In Figures 5C - D, the authors generated and used new lines for labeling pC1a and pC1b+c. The line for pC1b+c was imaged as part of another recent study (https://doi.org/10.1073/pnas.2310841121). However, similar additional images of the pC1a line (i.e. 40x magnification and VNC expression) would be helpful in order to validate its specificity.

We have included the high-resolution images of the expression of the pC1a-split-Gal4 driver in the brain and the VNC in the new figures S6A and S6B.

(2) The author's experiments examining olfactory and gustatory contributions to the holding period were well controlled and described. However, the experiments in Figure 1D examining visual contributions were not sufficiently convincing as the line used (w1118) has previously been shown to be visually impaired (Wehner et al., 1969; Kalmus 1948). Using another wild-type line would have improved the authors' claims.

It is evident that w1118 flies are visually impaired and are able to receive a limited amount of visual information in dim red light. Nevertheless, they are able to exhibit MIES phenotypes, which further supports the dispensability of visual information in MIES. In a 2024 study, Doubovetzky et al. (1) found that MIES in ninaB mutant females, which have defects in visual sensation, was not altered. This further corroborates our assertion that vision is likely to be of lesser importance than olfaction in MIES.

(3) When comparisons between more than 2 groups are shown as in Figures 1E, 3D, and 5E, the comparisons being made were not clear. Adding in the results of a nonparametric multiple comparisons test would help for the interpretation of these results.

We have revised figures 1E, 3D, 5E and the accompanying legends as suggested.

Reviewer #2 (Public Review):

The work by Yun et al. explores an important question related to post-copulatory sexual selection and sperm competition: Can females actively influence the outcome of insemination by a particular male by modulating the storage and ejection of transferred sperm in response to contextual sensory stimuli? The present work is exemplary for how the Drosophila model can give detailed insight into the basic mechanism of sexual plasticity, addressing the underlying neuronal circuits on a genetic, molecular, and cellular level.

Using the Drosophila model, the authors show that the presence of other males or mated females after mating shortens the ejaculate-holding period (EHP) of a female, i.e. the time she takes until she ejects the mating plug and unstored sperm. Through a series of thorough and systematic experiments involving the manipulation of olfactory and chemo-gustatory neurons and genes in combination with exposure to defined pheromones, they uncover two pheromones and their sensory cells for this behavior. Exposure to the male-specific pheromone 2MC shortens EHP via female Or47b olfactory neurons, and the contact pheromone 7-T, present in males and on mated females, does so via ppk23 expressing gustatory foreleg neurons. Both compounds increase cAMP levels in a specific subset of central brain receptivity circuit neurons, the pC1b,c neurons. By employing an optogenetically controlled adenyl cyclase, the authors show that increased cAMP levels in pC1b and c neurons increase their excitability upon male pheromone exposure, decrease female EHP, and increase the remating rate. This provides convincing evidence for the role of pC1b,c neurons in integrating information about the social environment and mediating not only virgin but also mated female post-copulatory mate choice.

Understanding context and state-dependent sexual behavior is of fundamental interest. Mate behavior is highly context-dependent. In animals subjected to sperm competition, the complexities of optimal mate choice have attracted a long history of sophisticated modelling in the framework of game theory. These models are in stark contrast to how little we understand so far about the biological and neurophysiological mechanisms of how females implement post-copulatory or so-called "cryptic" mate choice and bias sperm usage when mating multiple times.

The strength of the paper is decrypting "cryptic" mate choice, i.e. the clear identification of physiological mechanisms and proximal causes for female post-copulatory mate choice. The discovery of peripheral chemosensory nodes and neurophysiological mechanisms in central circuit nodes will provide a fruitful starting point to fully map the circuits for female receptivity and mate choice during the whole gamut of female life history.

We appreciate the positive response to our work.

Recommendations for the authors:

Reviewing Editor (Recommendations For The Authors):

While appreciating the quality of the work the reviewers had a few key concerns that would greatly improve the manuscript. These are:

(1) In some cases the specific statistical analyses are not clear. Could the authors please clarify what comparisons were made and the specific tests used?

We have clarified the comparisons made in the multiple comparison analysis and specified the tests used in figures 1E, 3D, 5E.

(2) Could the authors please include data that verify the expression patterns of their new reagent for pC1a, which will be useful for the community?

Figure S6 was revised to include the expression of the pC1a-split-Gal4 gene in the brain (Fig. S6A) and the VNC (Fig. S6B).

(3) A figure summarising their findings in the context of known circuitry will be useful.

A new Figure 7 has been prepared, which provides a summary of our findings.

(4) The SAG data are interesting. Do the authors wish to consider moving it to the main text or removing it if too preliminary?

The supplementary figure 10 and related discussions in the discussion section have been removed.

In the revised version of this manuscript, we present new evidence that the Or47b gene is required for 2MC-induced cAMP elevation in pC1 neurons, but not for 7T-induced one (see Fig. 5F). This observation supports that Or47b is a receptor for 2MC.

The following paragraph was inserted at line 248 to provide a detailed description of the new findings: "To further test the role of Or47b in 2MC detection, we generated Or47b-deficient females with pC1 neurons expressing the CRE-luciferase reporter. Females with one copy of the wild-type Or47b allele, which served as the control group, showed robust CRE-luciferase reporter activity in response to either 2MC or 7-T. In contrast, Or47b-deficient females showed robust CRE-luciferase activity in response to to 7-T, but little activity in response to 2MC. This observation suggests that the odorant receptor Or47b plays an essential role in the selective detection of 2MC (Fig. 5F).”

In addition, the following sentence was inserted at line 308 in the discussion section: “In this study, we provide compelling evidence that 2MC induces cAMP elevation in pC1 neurons and EHP shortening via both the Or47b receptor and Or47b ORNs, suggesting that 2MC functions as an odorant ligand for Or47b.”

Relative CRE-luciferase reporter activity of pC1 neurons in females of the indicated genotypes, incubated with a piece of filter paper perfumed with solvent vehicle control or the indicated pheromones immediately after mating. The CRE-luciferase reporter activity of pC1 neurons of Or47b-deficient females (Or47b2/2 or Or47b3/3) was observed to increase in response to 7-T but not to 2MC. To calculate the relative luciferase activity, the average luminescence unit values of the female incubated with the vehicle are set to 100%. Mann-Whitney Test (n.s. p > 0.05; **p < 0.01; ***p < 0.001; ****p < 0.0001). Gray circles indicate the relative luciferase activity (%) of individual females, and the mean ± SEM of data is presented.

Reviewer #1 (Recommendations For The Authors):

(1) There was a discrepancy between the text and the figures. Based on the asterisks above the data in Figure S5A, the data supports only 150 ng of 7-T shortening the ejaculation holding period. However, the text states that (line 190) "150 or 375 ng of 7-T significantly shortened EHP." It would be helpful if the authors clarified this discrepancy.

The sentence has been revised and now reads as follows: ‘150 ng of 7-T significantly shortened EHP’.

(2) Based on the current organization of the text, it was not clear how 2MC was identified and its concentrations were known to be physiologically relevant. It would be helpful if the authors could expand on this in lines 178 - 179.

The following sentences were inserted into the revised version of the manuscript at line 178: The EHP was therefore measured in females incubated in a small mating chamber containing a piece of filter paper perfumed with male CHCs, including 2-methylhexacosane, 2-methyldocosane, 5-methyltricosane, 7-methyltricosane, 10Z-heneicosene, 9Z-heneicosene, and 2MC at various concentrations (not shown). Among these, 2MC at 750 ng was the only one that significantly reduced EHP (Fig. 3A; Fig. S4). 2MC was mainly found in males, but not in virgin females (30). Notably, it is present in D. melanogaster, D. simulans, D. sechellia, and D. erecta, but not in D. yakuba (30, 60).

(3) The inset pie chart image illustrating MIES in Figure 1A was difficult to interpret. It would be helpful if the authors used a different method for representing this (i.e. a timeline).

Figure 1A was revised as suggested.

(4) In lines 121 - 122, the authors state that the females are exposed to "actively courting naive wild type Canton S males." This was difficult to understand and might be improved by removing "actively courting."

Revised as suggested.

Reviewer #2 (Recommendations For The Authors):

(1) Summary figure

The story is quite comprehensive and contains a lot of detail regarding the interaction of signaling pathways, internal state, and sensory stimuli. I believe a schematic summary figure bringing together all findings could be very helpful and would make it much easier to understand the discussion!

Figure 7 has been prepared, which provides a summary of the findings and an explanation of the current working model.

(2) Figure S10/effect on SAG activation of EHP

At the moment, the quite interesting and relevant result that SAG activation shortens EHP shown in Figure S10 is only referred to in the discussion. Maybe move this to the results and give it a bit more attention? Actually, I believe this is a very exciting finding that could also be the basis for some more interesting speculations about physiological relevance. Since SAG is silenced upon seminal fluid/sex peptide exposure after mating, a mating with failed SAG silencing (i.e. unusually high post-mating SAG activity) could indicate to the female that there was low or failed sex peptide/seminal fluid transfer. In such a case it would be probably advantageous for the female to decrease EHP and quickly remate, as females need the "beneficial" effects of seminal fluid on ovulation and physiology adaptation. SAG could therefore represent another arm of sensing male quality- here not via external pheromones, but internally, via sensing male sex peptide levels.

If this is a bit preliminary and rather suited to start a new study, Figure S10 could also be removed from the current manuscript.

Figure S10 and associated text were removed in the revised version of the manuscript.

(3) PhotoAC experiments in pC1b,c: the authors find that raising cAMP levels in pC1b,c leads to a decrease in EHP. They argue that increased cAMP levels lead to higher excitability of pC1b,c. This implies that the activity of pC1b,c promotes mating plug ejection. I assume the authors have also tried activating pC1b,c directly by optogenetic cation channels? What is the outcome of this? If different from elevating cAMP levels: why so?

We employed CsChrimson, a red light-sensitive channelrhodopsin, to investigate the effect of optogenetic activation of each pC1 subset on EHP. Optogenetic activation of pC1a, pC1d, or pC1e had little effect on EHP; however, optogenetic activation of pC1b, c significantly increased EHP. This observation was puzzling because optogenetic silencing of the same neurons also increased EHP. In this experiment, females expressing CsChrimson were exposed to red light for the entire period of EHP measurement. Therefore, we suspect that prolonged activation of pC1b and pC1c neurons depleted their neurotransmitter pool, resulting in a silencing effect, but this requires further testing.

The prolonged optogenetic activation of pC1b, c neurons increases EHP, mimicking silencing of pC1b, c neurons.

Females of the indicated genotypes were cultured on food with or without all-trans-retinal (ATR). The ΔEHP is calculated by subtracting the mean of the reference EHP of females cultured in control ATR- food from the EHP of individual females in comparison. The female genotypes are as follows: (A) 71G01-GAL4/UAS-CsChrimson, (B) pC1a-split-Gal4/UAS-CsChrimson, (C) pC1b,c-split-Gal4/UAS-CsChrimson, (D) pC1d-split-Gal4/UAS-CsChrimson, and (E) pC1e-split-Gal4/UAS-CsChrimson. Gray circles indicate the ΔEHP of individual females, and the mean ± SEM of data is presented. Mann-Whitney Test (n.s. p > 0.05; *p <0.05; ****p < 0.0001). Numbers below the horizontal bar represent the mean of the EHP differences between the indicated treatments.

(4) Text edits

In general, the manuscript is very well-written, clear, and easy to follow. I recommend small edits of the text and correction of typos in some places:

l.92: "Drosophila females seem to signal the social sexual context through sperm ejection." This sentence could give the impression that the main function of sperm ejection was to signal to conspecifics. I recommend reformulating to leave it open if ejected sperm is a signal or rather a simple cue. e.g. :"There is evidence that Drosophila females detect the social sexual context through sperm ejected by other females."

Thanks for the good suggestion. It has been revised as suggested. In addition, we have also made additional changes to the text to correct typos.

l.97: "transcriptional factor" > "transcription factor"

Revised as suggested. See lines 77, 98, and 201.

l.101: "There are Dsx positive 14 pC1 neurons in each brain hemisphere of the brain," > "There are 14 Dsx positive pC1 neurons in each brain hemisphere,"

Revised as suggested, it now reads " There are 14 Dsx-positive pC1 neurons in each hemisphere of the brain, ...".

l.160: ", even up to 1440 ng" > ", even when applied at concentrations as high as 1440 ng"

Revised as suggested.

l.168: "females with male oenocytes significantly shortens EHP" >"females with male oenocytes significantly shorten EHP"

Revised as suggested.

l.181: "it was restored when Orco expression is reinstated" >"it was restored when Orco expression was reinstated"

Revised as suggested. See line 186.

l.196: "MIES is almost completely abolished" >"MIES was almost completely abolished"

Revised as suggested. See line 201.

l.202: "a sexually dimorphic transcriptional factor gene" >"the sexually determination transcription factor gene" or "the sex specifically spliced transcription factor gene". The gene itself is not dimorphic!

Revised as suggested, lines 208-210 now read "The same study found that Dh44 receptor neurons involved in EHP regulation also express doublesex (dsx), which encodes sexually dimorphic transcription factors."

l.211: "to silenced" > "to silence"

Revised as suggested. See line 216.

l.229: "females that selectively produce the CRE-Luciferase reporter gene" >"females that selectively express CRE-Luciferase reporter"

Revised as suggested. See line 234.

l.271: "neurons. expedite" > delete dot

Revised as suggested. See line 284.

l.287: "Furthermore, our study has uncovered the conserved neural circuitry that processes male courtship cues and governs mating decisions play an important role in regulating this behavior." > grammar: "our study has uncovered that the conserved neural circuitry that processes male courtship cues and governs mating decisions plays an important role in regulating this behavior." Also: the meaning of "conserved" is not fully clear to me here: conserved in regards to other Drosophila species? Or do the authors mean: general functional similarity with mouse sexual circuitry?

The sentence (lines 299-301) has been revised for clarity to read "In addition, our study has revealed that the neural circuit that processes male courtship cues and controls mating decisions plays an important role in regulating this behavior. This fly circuit has recently been proposed to be homologous to VMHvl in the mouse brain (45, 46).”

l.311: "lipid drolet" > "lipid droplets"

Revised as suggested. See line 325.

l.316 and in several instances in the following, including Figure 5 caption (l.723) : "cAMP activity" > "cAMP levels" or "increased cAMP levels"

Revised as suggested.

l.323: "in hemibrain" > ", as seen in the hemibrain connectome dataset"

Revised as suggested. See line 337.

l.326: "increased cAMP levels causes pC1b,c neurons" > "increased cAMP levels cause pC1b,c neurons"

Revised as suggested. See line 340.

l.329: "removement" > "removal" or "ejection"

Revised as suggested, it now reads "the removal of the mating plug". See line 343.

l. 330: "This observation well aligns" > "The observation aligns well"

Revised as suggested. See line 345.

l. 398: Behavior assays: It would be good to describe how mating plug ejection was identified- by eye? Under the microscope/UV light?

The following sentence has been added to the behavioral assays section at lines 425-426: The sperm ejection scene, in which the female expels a white sac containing sperm and the mating plug through the vulva, has been directly observed by eye in recorded video footage.

l.685, Figure legend 2: "thermal activation" > "thermogenetic activation"

Revised as suggested. See line 430.

Reference:

(1) Doubovetzky, N., Kohlmeier, P., Bal, S., & Billeter, J. C. (2023). Cryptic female choice in response to male pheromones in Drosophila melanogaster. bioRxiv, 2023-12.

Version 2

Abstract

Abstract

In birds and insects, the female uptakes sperm for a specific duration post-copulation known as the ejaculate holding period (EHP) before expelling unused sperm and the mating plug through sperm ejection. In this study, we found that Drosophila melanogaster females shortens the EHP when incubated with males or mated females shortly after the first mating. This phenomenon, which we termed male-induced EHP shortening (MIES), requires Or47b+ olfactory and ppk23+ gustatory neurons, activated by 2-methyltetracosane and 7-tricosene, respectively. These odorants raise cAMP levels in pC1 neurons, responsible for processing male courtship cues and regulating female mating receptivity. Elevated cAMP levels in pC1 neurons reduce EHP and reinstate their responsiveness to male courtship cues, promoting re-mating with faster sperm ejection. This study established MIES as a genetically tractable model of sexual plasticity with a conserved neural mechanism.

Introduction

Sexual plasticity, the ability to modify sexual state or reproductive behavior in response to changing social conditions, is observed in both vertebrates and invertebrates (1–5). In rodents, exposure to unfamiliar males often leads to the sudden termination of pregnancy, known as the Bruce effect. It is induced by male urinary peptides, such as MHC I peptides, activating the vomeronasal organ (6–8). This effect enhances reproductive fitness of both sexes, by eliminating the offspring of competing males and enabling females to select better mates even after conception. Many species also adapt their reproductive behavior in response to the social sexual context change (SSCC), involving encounters with new sexual partners or competitors. Understanding the neural circuit mechanisms behind female responses to SSCC emerges as a central focus of neuroscience (9–12).

Drosophila melanogaster, the fruit fly, displays various social behaviors like aggregation, aggression, and sexual behavior (13–15). Similar to rodents, they primarily use the olfactory system to communicate socially through pheromones (16, 17). Some of these pheromones act as aphrodisiacs, while others regulate aggression or foster aggregation. For instance, cis-vaccenyl acetate (cVA) attracts females but repels males and promotes aggregation (14, 18, 19). 7-Tricosene (7-T), a cuticular hydrocarbon (CHC) present in males, is an aphrodisiac to females and affects social interactions between males (20, 21). On the other hand, 7,11-heptacosadiene (7,11-HD), a related female-specific pheromone, functions as an aphrodisiac to males, triggering courtship behavior and involving species recognition (22, 23).

The fruit fly’s chemo-sensory organs, located in various parts of the body, detect these pheromones (24, 25). Olfactory receptor neurons (ORNs) in the sensilla of the antennae and maxillary palp are responsible for the detection of long-range volatile pheromones like cVA, while short-range pheromones like 7-T are sensed by neurons on the fore-legs and labellum (16, 17, 24).

The olfactory receptor Or47b, expressed ORNs located in at4 trichoid sensilla on the third antennal segment, is involved in several socio-sexual interactions, including male mating success, mate preference, and female aggression toward mating pairs (12, 26–29). In males, Or47b senses fatty acid methyl esters and fatty acids that affect mating competition and copulation (30, 31). While the role of Or47b in female aggression is well established (12), its involvement in female sexual behavior is uncertain. In both sexes, Or47b ORNs project to VA1v glomeruli, where VA1v projection neurons receive their signal and project to the mushroom body calyx and lateral horn. Male Or47b neurons connect to neurons such as aSP5, aSP8, and aSP9, which express a male-specific transcription factor FruM (32).

CHC pheromones, which function as short-range pheromones, are detected primarily by neurons on the fore-legs and the labellum that express gustatory receptors (GR), ionotropic receptors (IR), or the ppk/DEG-ENaC family of sodium channels (16, 17, 24). CHCs like 7-T and 7,11-HD are sensed by ppk23-expressing M and F cells in the tarsi (33). 7-T and cVA are sensed by M cells expressing ppk23, whereas 7,11-HD and 7,11-nonacosadiene (7,11-ND) are sensed by F cells expressing ppk23, ppk25, and ppk29. In males 7-T or 7,11-HD affects the neuronal activity of the FruM-expressing P1 neurons (34–36). However, how these CHCs signal in the female brain remains unknown.

Sperm ejection is a process by which females can remove the male ejaculate or the mating plug after copulation. This phenomenon has been observed in several animal species including feral fowl (37), black-legged kittiwake (38), and dunnock (39). In the fruit fly, it typically occurs approximately 90 minutes after mating (40). This specific interval, referred to as ejaculate holding period (EHP), is thought to affect sperm usage and fecundity (40, 41). The neurosecretory neurons in the brain pars intercerebralis (PI) that produce diuretic hormone 44 (Dh44), an insect orthologue of the corticotropin-releasing factor, regulate EHP (40). There is evidence that Drosophila females sense the social-sexual context through sperm ejected by other females. For instance, females were likely to lay more eggs when placed on a food patch containing male ejaculate deposited by other females (42). However, it remains unknown whether the SSCC influences sperm ejection and EHP.

Female pC1 neurons, which express a specific transcription factor DsxF, integrate olfactory and auditory cues associated with male courtship (43, 44). The pC1 neurons, their male counterparts (i.e., P1 neurons), and the ventrolateral subdivision of ventromedial hypothalamus (VMHvl) neurons in mice share conserved circuit configurations and demonstrate functional similarity in coordinating social and sexual behaviors (45, 46). There are 14 Dsx-positive pC1 neurons in each hemisphere of the brain, responsive to the male sex-pheromone cVA and courtship songs (44, 47). Connectome analyses identified 10 pC1 neurons that fall into five subtypes, with pC1a, b, and c subtypes associated with mating behavior and pC1d and e subtypes associated with aggression (47–52). Although direct evidence connecting pC1 neurons to sperm ejection is limited, they are promising candidates for regulating sperm ejection or EHP, because sperm ejection allows females to eliminate the mating plug and male ejaculate, thereby restoring sexual attractiveness (53).

In this study, we demonstrated that two male pheromones, 2-methyltetracosane (2MC) and 7-T, significantly reduced the EHP through Or47b neurons and ppk23 neurons, respectively. These pheromone pathways converge on pC1 neurons, where they increase cAMP levels. The elevated cAMP in pC1 neurons resulted in a reduction of the EHP to a degree that was comparable to the effects of the male pheromones. It also enhanced the excitability of pC1 neurons, making them more responsive to both olfactory and auditory male courtship cues and promoting further mating following the earlier removal of the mating plug. These findings establish a novel behavioral paradigm that sheds light on the intricate molecular and neuronal pathways underlying female sexual plasticity.

Results

Male-induced EHP shortening (MIES) is dependent on olfaction

To investigate the impact of changes in the social sexual context on the EHP, we compared the EHP of post-mating females isolated from any male presence to those exposed to naive wild-type Canton-S (CS) males immediately after copulation (Fig. 1A). Notably, the EHP of females incubated with naive males was approximately 30 minutes shorter than that of females left in isolation after mating (Fig. 1A, 1B). We refer to this phenomenon as male-induced EHP shortening (MIES). In contrast, little difference in EHP was observed between females incubated with virgin females and those isolated after mating (Fig. 1C).

The presence of males reduces the ejaculate holding period (EHP) in females through olfactory or gustatory sensation

A, Schematic of the experimental procedure employed to measure male-induced EHP shortening (MIES). Immediately after the end of copulation, the female is incubated with a wild-type Canton-S (CS) male that has not been previously exposed to the female. Typically, w1118females that are kept alone after mating exhibit an EHP of approximately 90 minutes, whereas females that are incubated with a naïve CS male exhibit an EHP of approximately 60 minutes. In this study, we refer to this phenomenon as MIES. B-F, Normalized ejaculate holding period (EHP) or ΔEHP of the females of the indicated genotypes, incubated under the indicated conditions after mating. The ΔEHP is calculated by subtracting the mean of the reference EHP of females kept alone after mating (the leftmost column) from the EHP of individual females in comparison. Mann-Whitney Test (n.s. p > 0.05; ****p < 0.0001). Gray circles indicate the EHP or ΔEHP of individual females, and the mean ± SEM of data is presented. Numbers below the horizontal bar represent the mean of the EHP differences between the indicated treatments.

Male fruit flies employ various sensory signals to attract females during courtship (15). To assess the role of the visual signal in MIES, we examined MIES under dim red light conditions and observed that limited illumination had a marginal impact on MIES (Fig. 1D). Next, we examined MIES in post-mating females incubated with decapitated CS males. These males could serve as a source of olfactory or gustatory signals, but not for auditory or visual signals. Again, no reduction in MIES was observed (Fig. 1E). This strongly suggests that olfactory or gustatory cues are the key signals responsible for MIES. This is further supported by the observation that females deficient in the odorant receptor co-receptor (Orco1) did not exhibit MIES (Fig. 1F). Thus, it is highly likely that male odorant(s), especially those detected by olfactory receptors (Or), induce MIES.

MIES is dependent on the Or47b receptor and Or47b-expressing ORNs

In the fruit fly antenna, the trichoid sensilla and their associated olfactory receptor neurons are known to detect sex pheromones (54). To investigate the contribution of ORNs located in the trichoid sensilla to MIES, we silenced 11 different ORN groups found in the trichoid and intermediate sensilla (55, 56) by expressing either the active or inactive form of Tetanus toxin light chain (TNT) (57). Our results showed that silencing ORNs expressing Or13a, Or19a, Or23a, Or47b, Or65c, Or67d or Or88a significantly affected MIES (Fig. S1A).

We then focused on the analysis of Or47b-positive ORNs (Fig. 2A), which, in contrast to the others, exhibited almost complete abolition of MIES when silenced. Activation of these neurons with the thermogenetic activator dTRPA1 (58) resulted in a significant EHP shortening, even in the absence of male exposure (Fig. 2B). Subsequently, we examined whether restoring Orco expression in Or47b ORNs in Orco-deficient females would restore MIES. Our results confirmed that this is indeed the case (Fig. 2C). To establish the necessity of the Or47b receptor gene for MIES, we examined Or47b-deficient females (Or47b2/Or47b3) and observed a complete absence of MIES, whereas heterozygous controls exhibited normal MIES (Fig. 2D). Furthermore, the reintroduction of Or47b expression in Or47b ORNs of Or47b-deficient females almost completely restored MIES (Fig. 2E). Based on these observations, we concluded that MIES depends on the Or47b receptor gene and Or47b-expressing ORNs.

The function of Or47b and Or47b-positive ORNs is essential for MIES

A, C-E, ΔEHP of females of the indicated genotypes, incubated with or without naive males after mating. The female genotypes are as follows from left to right: (A) control (Or47b>TNTinactive), Or47b ORN silencing (Or47b>TNTactive); (C) Orco mutant (Orco1/Orco1), Orco rescue in Or47b ORNs of Orco mutant (Orco1/Orco1; Or47b>Orco); (D) control 1 (Or47b2/+), control 2 (Or47b3/+), Or47b mutant (Or47b2/Or47b3); (E) Or47b mutant (Or47b2/Or47b2), Or47b rescue (Or47b>Or47b; Or47b2/Or47b2).

B, Thermogenetic activation of Or47b-positive ORNs shortens EHP in females kept alone after mating. The female genotypes are as follows from left to right: control 1 (Or47b-Gal4/+), control 2 (UAS-dTRPA1/+), Or47b>dTRPA1 (Or47b-Gal4/UAS-dTRPA1).

Mann-Whitney Test (n.s. p > 0.05; *p <0.05; **p <0.01; ****p < 0.0001). The ΔEHP is calculated by subtracting the mean of the reference EHP of females kept alone after mating (‘-’ in A, C-E) or incubated at 21°C control conditions (B) from the EHP of individual females in comparison. Gray circles indicate the ΔEHP of individual females, and the mean ± SEM of data is presented. The gray circles with dashed borders indicate ΔEHP values that exceed the axis limits (>90 or <-90 minutes). Numbers below the horizontal bar represent the mean of the EHP differences between the indicated treatments.

2-Methyltetracosane (2MC) induces MIES via Or47b and Or47b ORNs

Previous studies have shown that methyl laurate (ML) and palmitoleic acid (PA) can activate Or47b ORNs only in the presence of a functional Or47b gene (30, 31). However, in our investigation, none of these odorants induced significant EHP shortening, even when applied at concentrations as high as 1440 ng (Fig. S2). This prompted us to search for a new pheromone capable of activating Or47b ORNs and thereby shortening the EHP.

Oenocytes produce a significant portion of the cuticular hydrocarbons or pheromones. We asked whether the male pheromone responsible for MIES is produced by oenocytes (Fig. S3A). Indeed, incubation with females engineered to produce male oenocytes significantly shorten EHP, strongly suggesting that male oenocytes serve as a source for the MIES pheromone. Unexpectedly, however, incubation with males possessing feminized oenocytes also resulted in significant EHP shortening (Fig. S3A). This raises the possibility that oenocytes may not be the sole source of the MIES pheromone, implying the involvement of more than one pheromone, for instance one from oenocytes and another from an alternate source, in MIES.

The genus Drosophila exhibits distinct CHC profiles, with certain CHC components shared among closely related species (59). We found that incubation with males of other closely related species, such as D. simulans, D. sechellia, and D. erecta, also induced EHP shortening, whereas incubation with D. yakuba males did not (Fig. S3B).

The EHP was therefore measured in females incubated in a small mating chamber containing a piece of filter paper perfumed with male CHCs, including 2-methylhexacosane, 2-methyldocosane, 5-methyltricosane, 7-methyltricosane, 10Z-heneicosene, 9Z-heneicosene, and 2MC at various concentrations (not shown). Among these, 2MC at 750 ng was the only one that significantly reduced EHP (Fig. 3A; Fig. S4). 2MC was mainly found in males, but not in virgin females (30). Notably, it is present in D. melanogaster, D. simulans, D. sechellia, and D. erecta, but not in D. yakuba (30, 60).

2-Methyltetracosane (2MC) can induce EHP shortening through Or47b

A-D, Δ EHP of females of the indicated genotypes, incubated in solvent vehicle or 2MC. Mated females were incubated with a piece of filter paper perfumed with either vehicle (-) or 750 ng 2MC (+). The female genotypes are as follows: (A) w1118, (B) Orco mutant (Orco1/Orco1), (C) Or47b mutant (Or47b2/Or47b2), (D) Gal4 control (Or47b-Gal4/+; Orco1/Orco1), UAS control (UAS-Orco/+; Orco1/Orco1), Orco rescue in Or47b ORN (Orco1/Orco1; Or47b-Gal4/UAS-Orco).

A-C , Mann-Whitney Test (n.s. p > 0.05; *p <0.05). D, One-way analysis of variance (ANOVA) test with Fisher’s LSD multiple comparison (n.s. p > 0.05; *p < 0.05). Gray circles indicate the ΔEHP of individual females and the mean ± SEM of data is presented. The ΔEHP is calculated by subtracting the mean of the reference EHP of females incubated with vehicle-perfumed paper (the leftmost column in A-C) or the mean of the Gal4 control and UAS control female incubated with vehicle-perfumed paper (the two leftmost columns in D) from the EHP of individual females in comparison. Gray circles with dashed borders indicate ΔEHP values that exceed the axis limits (>90 or <-90 minutes). Numbers below the horizontal bar represent the mean of the EHP differences between the indicated treatments.

Moreover, the 2MC-induced EHP shortening was not observed in Orco- or Or47b-deficient females (Fig. 3B, 3C), but was restored when Orco expression was reinstated in Or47b ORNs in Orco-deficient mutants (Fig. 3D). Our behavioral observations strongly suggest that 2MC acts as an odorant ligand for Or47b and shortens the EHP through this receptor.

7-Tricosene (7-T) shortens EHP through ppk23 neurons

In contrast to incubation with virgin females, incubation with mated females resulted in a significant shortening of EHP (Fig. 4A). Mated females carry male pheromones, including 7-T and cVA, which are transferred during copulation (53). This raised the possibility that these male pheromones might also induce EHP shortening. Indeed, our experiments revealed that incubation with a piece of filter paper perfumed with 150 ng of 7-T significantly shortened the EHP. Conversely, incubation with cVA and 7-pentacosene, a related CHC, did not produce the same effect (Fig. 4B, 4C; Fig. S5A-5B). The concentrations of 7-T capable of inducing EHP shortening appear to be physiologically relevant. 7-T has been found at levels of 432 ng in males (61), 25 ng in virgin females, and 150 ng in mated females (53). Although the receptors for 7-T remain unknown, ppk23-expressing tarsal neurons have been shown to sense these compounds and regulate sexual behavior in males and females (23, 62, 63). Subsequently, we silenced ppk23 neurons, and as a result, MIES was almost completely abolished, underscoring the pivotal role of 7-T in MIES (Fig. 4D). However, DEG/ENac channel genes expressed in ppk23 neurons, including ppk23 and ppk29, were found to be dispensable for MIES (Fig. S5C-5E). This aligns with previous observations that neither ppk23 deficiency nor ppk28 deficiency recapitulates the sexual behavioral defects caused by silencing ppk23 neurons (64).

7-Tricosene present in mated females and males reduces EHP via ppk23 neurons

A-D , ΔEHP of females of the indicated genotypes, incubated with mated females (A), a piece of filter paper perfumed with 150 ng 7-T (B), 200 ng cVA (C), or naive males (D) after mating. The female genotypes are as follows: (A-C) w1118, (D) control (ppk23-Gal4/UAS-TNTinactive), ppk23 silencing (ppk23-Gal4/UAS-TNTactive). A, Unpaired t-Test, B-D, Mann-Whitney Test (n.s. p > 0.05; *p < 0.05). The ΔEHP is calculated by subtracting the mean of the reference EHP of females kept alone (‘-’ in A, D) or incubated with vehicle-perfumed paper (the leftmost column in B, C) from the EHP of individual females in comparison. Gray circles indicate the ΔEHP of individual females, and the mean ± SEM of data is presented. The gray circles with dashed borders indicate ΔEHP values that exceed the axis limits (>90 or <-90 minutes). Numbers below the horizontal bar represent the mean of the EHP differences between the indicated treatments.

The pC1 b and c neurons regulate EHP and MIES

The neuropeptide Dh44 determines the timing of sperm ejection or EHP (40). The same study found that Dh44 receptor neurons involved in EHP regulation also express doublesex (dsx), which encodes sexually dimorphic transcription factors. A recent study has revealed that pC1 neurons, a specific subgroup of dsx-expressing central neurons in the female brain, do indeed express Dh44 receptors (47). With these findings, we set out to investigate the role of pC1 neurons in the regulation of EHP and MIES. The pC1 neurons comprise five distinct subtypes. Of these, the pC1a, b, and c subtypes have been implicated in mating receptivity (47, 48), while the remaining pC1d and e subtypes have been implicated in female aggression (48, 52). To investigate the role of these subtypes in EHP, we employed GtACR1, an anion channel activated by blue light in the presence of all trans-retinal (ATR), to silence specific pC1 subtypes immediately after mating. Our experiments revealed that silencing of the pC1 subset comprising the pC1a, b and c subtypes with GtACR1 led to an increase in EHP (Fig. 5A), whereas silencing of the pC1d and e subtypes had a limited effect on EHP (Fig. 5B). We further analyzed the roles of pC1b, c neurons along with pC1a neurons separately. We generated a subtype-specific split-Gal4 for pC1a and found that, as expected, silencing pC1a with this split-Gal4 almost completely suppressed mating receptivity (Fig. S6).

A subset of pC1 neurons, comprising pC1b and pC1c subtypes, regulates EHP and exhibits CRE-luciferase reporter activity in response to 2MC and 7-T

A-D, The optogenetic silencing of a pC1 neuron subset comprising pC1b and pC1c neurons (pC1b, c) increases EHP. Females of the indicated genotypes were cultured on food with or without all trans-retinal (ATR) after eclosion. The ΔEHP is calculated by subtracting the mean of the reference EHP of females cultured in control ATR-food (the leftmost column) from the EHP of individual females in comparison. The female genotypes are as follows: (A) pC1a,b,c>GtACR1 (pC1-S-Gal4/UAS-GtACR1), (B) pC1d,e>GtACR1 (pC1-A-Gal4/UAS-GtACR1), (C) pC1a>GtACR1 (pC1a-split-Gal4 /UAS-GtACR1), and (D) pC1b,c>GtACR1 (Dh44-pC1-Gal4/UAS-GtACR1). Gray circles indicate the ΔEHP of individual females, and the mean ± SEM of data is presented. The gray circles with dashed borders indicate ΔEHP values that exceed the axis limits (>120 minutes). Mann-Whitney Test (n.s. p > 0.05; *p <0.05; ****p < 0.0001).

Numbers below the horizontal bar represent the mean of the EHP differences between the indicated treatments.

E, F, Relative CRE-luciferase reporter activity of pC1 neurons in females of the indicated genotypes, incubated with a piece of filter paper perfumed with solvent vehicle control or the indicated pheromones immediately after mating. The CRE-luciferase reporter activity of pC1 neurons of Or47b-deficient females (Or47b2/2 or Or47b3/3) was observed to increase in response to 7-T but not to 2MC. To calculate the relative luciferase activity, the average luminescence unit values of the female incubated with the vehicle are set to 100%. Mann-Whitney Test (n.s. p > 0.05; **p < 0.01; ***p < 0.001; ****p < 0.0001). Gray circles indicate the relative luciferase activity (%) of individual females, and the mean ± SEM of data is presented.

However, silencing pC1a alone did not result in increased EHP, suggesting a marginal role of the pC1a subtype in EHP regulation (Fig. 5C). In contrast, concomitant silencing of both pC1b and pC1c neurons significantly increased EHP by 56 + 6.9 minutes (Fig. 5D). At present, we lack the genetic tools to further distinguish the roles of pC1b and pC1c subtypes separately.

2MC and 7-T increase cAMP levels in pC1 b and c neurons

Our recent research has shown that pC1 neurons exhibit elevated cAMP levels during sexual maturation, and that this increase in cAMP is closely related to heightened excitability of pC1 neurons (47). The same study also showed that a mating signal (i.e., sex peptide in the male seminal fluid) reduces cAMP levels in pC1 neurons. Thus, we hypothesized that male odorants responsible for inducing MIES, such as 2MC or 7-T, would elevate cAMP levels in pC1b, c neurons in newly mated females. This, in turn, would lead to increased excitability of pC1 neurons and, as a consequence, a reduction in the EHP. To monitor cAMP levels in these neurons, we prepared females that express a CRE-luciferase reporter selectively in pC1b, c neurons. Indeed, when exposed to 2MC or 7-T, pC1b, c neurons exhibited a significant increase in CRE-luciferase activity, indicating that these neurons produce higher levels of cAMP in response to these odorants (Fig. 5E). Notably, CRE-luciferase activity appeared to peak at specific odorant concentrations that induced significant shortening of the EHP (Fig. S7).

In contrast, when we examined other pC1 subsets, such as pC1a, and pC1d and e, we detected no evidence of increased CRE-luciferase reporter activity upon exposure to 2MC or 7-T treatment (Fig. 5E). Notably, CRE-luciferase reporter activity in the pC1a neurons appears to be dependent on the mating status, as it reaches levels similar to those of pC1b, c neurons in virgin females (Fig. S8). This observation aligns well with connectome data suggesting that SAG neurons, which are responsible for relaying SP-dependent mating signals, synapse primarily with the pC1a subtype and to a much lesser extent with other pC1 subtypes (49).

To further test the role of Or47b in 2MC detection, we generated Or47b-deficient females with pC1 neurons expressing the CRE-luciferase reporter. Females with one copy of the wild-type Or47b allele, which served as the control group, showed robust CRE-luciferase reporter activity in response to either 2MC or 7-T. In contrast, Or47b-deficient females showed robust CRE-luciferase activity in response to to 7-T, but little activity in response to 2MC. This observation suggests that the odorant receptor Or47b plays an essential role in the selective detection of 2MC (Fig. 5F).

Elevated cAMP in pC1 neurons shortens the EHP, while increasing re-mating

Having shown that MIES-inducing male odorants, 2MC or 7-T, increase cAMP levels in pC1b, c neurons from mated females, we next asked whether this induced elevation of cAMP levels in pC1b, c neurons would shorten EHP, leading to MIES. We employed the photoactivatable adenylate cyclase (PhotoAC), which increases cellular cAMP levels upon exposure to light. Indeed, the induced elevation of cAMP levels in pC1b, c neurons significantly shortened EHP, whereas the same treatment applied to pC1a or pC1d and pC1e had no such effect (Fig. 6A). This further underscores the pivotal role of pC1b, c neurons in EHP regulation.

Elevated cAMP levels in pC1 neurons reduce EHP and increase the responsiveness of pC1 neurons to male courtship cues, thereby promoting subsequent remating

A, The optogenetic production of cAMP in the pC1b, c neurons shortens EHP, whereas the same treatment in pC1a or pC1d, e neurons does not. ΔEHP is calculated by subtracting the mean of the reference EHP of females incubated in the control illumination (Dim light), which does not activate a photoactivatable adenylate cyclase (PhotoAC), from the EHP of individual females. Mann-Whitney Test (n.s. p > 0.05, ****p < 0.0001).

B, The optogenetic production of cAMP transiently increases the excitability of pC1 neurons. Left, schematic of the experimental procedure. Right, peak ΔF/F in the LPC projections of pC1 neurons from freshly mated females in response to the pheromone cVA, before and after photoactivation of PhotoAC expressed in pC1 neurons. The calcium response was measured at specific time points after photoactivation: after 1 minute (blue dots and box) or 10 minutes (purple dots and box) after activation. Repeated measures one-way ANOVA test with the Geisser-Greenhouse correction followed by Tukey’s multiple comparisons test (*p < 0.05; ***p < 0.001; ****p < 0.0001).

C, Left, schematic of the experimental procedure. Right, re-mating rate of females during optogenetic cAMP production in pC1b, c neurons, scored as the percentage of females that copulate with a naive CS male within 6 hours after the first mating. The female genotypes are as follows: Control (+/UAS-PhotoAC), pC1b,c>UAS-PhotoAC (Dh44-pC1-Gal4/UAS-PhotoAC). Chi-square test (*p <0.05).

Next, we asked whether the expression of Dh44R1 and Dh44R2, GPCRs that increase cellular cAMP in response to their ligand Dh44, in pC1b, c neurons is necessary for MIES. However, double knockdown of Dh44R1 and Dh44R2 in pC1 neurons seemed to have a limited impact on MIES (Fig. S9). This suggests that Dh44R signaling in pC1 neurons is not essential for the regulation of EHP or MIES, raising the possibility that other GPCRs may be involved in the up-regulation of cAMP levels in pC1 neurons in response to 2MC or 7-T.

Lastly, we investigated how increased cAMP levels affect the physiological activity of pC1 neurons. pC1 neurons from virgin females exhibit robust Ca2+ transients in response to male courtship cues, such as the male pheromone cVA and the courtship pulse song (44). In contrast, those from mated females display significantly diminished Ca2+ transients (47). When examined shortly after mating, a decrease in pC1 responsiveness to cVA was observed. However, immediately after PhotoAC activation in pC1 neurons, pC1 neurons from freshly mated females became more excitable and exhibited stronger Ca2+ transients in response to cVA (Fig. 6B). It is important to note that this PhotoAC-induced increase in pC1 excitability is transient and rapidly declines within 10 minutes (Fig. 6B). Nevertheless, these findings suggest that the increased cAMP levels in pC1 neurons would not only promote MIES but also facilitate re-mating in post-mating females, which typically engage in re-mating at a low frequency. To test this hypothesis, we examined the re-mating frequency of freshly mated females paired with naive males while inducing a cAMP increase in pC1 neurons. As expected, PhotoAC activation in pC1b, c neurons substantially increased the re-mating rate compared to the control group (Fig. 6C). Therefore, we concluded that male odorants that stimulate cAMP elevation in pC1 neurons expedite the removal of the mating plug, consequently leading to increased instances of re-mating (Fig. 7).

The presence of male odorants, which reflect changes in the social-sexual context, stimulates newly mated females to remove the male ejaculate and engage in subsequent re-mating

Following the initial mating, a female that encounters a new courting male removes the male ejaculate after a shorter EHP than those that do not encounter new male partners. This phenomenon, referred to as MIES in this study, is followed by a second mating with the new partner. The production of MIES depends on the functions of the Or47b+ olfactory and ppk23+ gustatory neurons, which are activated by 2MC and 7-T, respectively. These odorants increase cAMP levels in pC1b, c neurons, enhancing their responsiveness to male courtship cues and increasing mating receptivity. Consequently, 2MC and 7-T promote a second mating with a faster removal of the male ejaculate or mating plug.

Discussion

Males employ a diverse range of strategies to enhance their reproductive fitness. One such strategy involves the formation ‘f a ’mating’plug’, a mechanism that prevents females from engaging in further mating, thereby increasing fertilization success rates (65–67). As a means of intra-sexual competition, rival males often promote the removal or precocious expulsion of the mating plug. The evolution of this strategy is driven by intersexual interactions with polyandrous females, who often remove the mating plug to engage in additional mating with males of superior traits or higher social status than their previous partners (37, 68). In the dunnock Prunella modularis, a small European passerine bird, the male often engages in cloacal pecking of mated females, inducing the expulsion of the previous ’mate’s sperm and mating plug, thereby increasing their chances of successful mating (39). In this study, we discovered that in D. melanogaster, freshly mated females exhibit an earlier removal of the mating plugs or a shorter EHP when kept with actively courting males. This behavior is primarily induced by the stimulation of females via male sex pheromones. In addition, our study has revealed that the neural circuit that processes male courtship cues and controls mating decisions plays an important role in regulating this behavior. This fly circuit has recently been proposed to be homologous to VMHvl in the mouse brain (45, 46). By delving into the molecular and neuronal mechanisms underlying MIES, our study provides valuable insights into the broader aspect of behaviors induced by changes in the social sexual context.

Our findings highlight the involvement of the Or47b receptor and Or47b ORNs in MIES. These OR and ORNs have been implicated in a range of social and sexual behaviors in both male and female fruit flies (12, 26–31). Methyl laurate and trans-palmitoleic acid are odorant ligands for Or47b that account for many of these functions particularly in males (30, 31). In this study, we provide compelling evidence that 2MC induces cAMP elevation in pC1 neurons and EHP shortening via both the Or47b receptor and Or47b ORNs, suggesting that 2MC functions as an odorant ligand for Or47b. Notably, gas chromatography–mass spectrometry (GC-MS) analysis of cuticular hydrocarbons from 4-day-old wild-type D. melanogaster revealed the presence of 2MC exclusively in males (30). Surprisingly, however, unlike 2MC, neither methyl laurate nor trans-palmitoleic acid affected EHP. The reason for this paradoxical result remains unclear. A plausible interpretation is that the EHP shortening induced by 2MC may require not only Or47b but also other as yet unidentified ORs. With the establishment of a behavioral and cellular assessment of 2MC activity, the search for additional odorant receptors responsive to 2MC is now feasible. Another important avenue for further research is whether 2MC can also elicit behaviors previously associated with methyl laurate or trans-palmitoleic acid, such as promoting male copulation and courtship (30, 31).

We observed that both 2MC and 7-T exhibit both cellular and behavioral activity within a specific concentration range (Fig. S4, S5, S7). This observation is of particular interest, given the multitude of environmental and biological factors that influence the levels of 2MC and 7-T, potentially affecting the capacity of males to induce MIES. For instance, exposure to low temperatures during development has been linked to increased production of both 2MC and 7-T (69). Similarly, mutation of the desiccation stress gene CG9186, which encodes a protein associated with lipid droplets, has been found to impact 2MC levels (70). Furthermore, 2MC levels rise with age in males (71). Thus, we propose that levels of 2MC, and possibly 7-T, may serve as indicators of male age and resilience to environmental stress in a complex manner.

In mated females, treatment with 2MC or 7-T increases cAMP levels in pC1b, c neurons but not in pC1a neurons. In contrast, pC1a neurons in virgin females are fully responsive to both male pheromones, showing increases in cAMP levels that are similar to those of pC1b,c neurons (Fig. S8). The absence of cAMP levels in pC1a neurons in mated females likely results from the mating signal (i.e., sex peptide) silencing pC1a neurons. Connectome and electrophysiology data support this interpretation, as SAG neurons, which relay sex peptide signals, have the strongest synaptic connection with the pC1a among five pC1 subtypes (49). However, the activity of SAG neurons may also influence pC1c neurons, as they also have substantial synaptic connections with pC1c neurons, as seen in the hemibrain connectome dataset (72). Future studies are needed to understand the role of SAG neurons in the regulation of EHP.

We found that increased cAMP levels cause pC1b, c neurons in mated females, which are typically unresponsive to male courtship cues like cVA and pulse song, to become responsive and exhibit strong Ca2+ transients. Since pC1b, c neurons play a role in generating sexual drive and increasing female receptivity to male courtship, the 2MC- or 7-T-induced increases in cAMP are likely to control the removal of the mating plug and the engagement of mated females in further mating (Fig. 7). This hypothesis aligns well with the previous report that mating reduces the sensitivity of Or47b ORNs, which we found to be responsive to 2MC, leading to an increased preference for pheromone-rich males after mating (29).

Moreover, the finding that 2MC and 7-T induce cAMP levels in pC1b, c neurons in virgin females suggests that virgin females may also use 2MC and 7-T as odorant cues to assess male quality during their first mating. Indeed, females seem to evaluate male quality by the amount of 7-T, as increased 7-T promotes mating receptivity and shortens mating latency (20).

Physiological factors like the nutritional status of females prior to mating and the nutritional status of their mates have been shown to influence EHP (73), and therefore potentially MIES. Hence, it is highly probable that MIES is regulated by additional central neurons such as Dh44-PI neurons that regulate these processes (40). However, it remains unclear whether and how Dh44-PI neurons and pC1 neurons interact to modulate EHP and MIES. The observation that double knockdown of Dh44R1 and Dh44R2 has only a marginal effect on MIES suggests that Dh44-PI neurons may also function independently of pC1 neurons, raising the possibility that multiple independent central circuits may contribute to the production of MIES.

Our initial screening of ORNs responsible for MIES revealed the involvement of Or47b ORNs, as well as several other ORNs. In addition to 2MC, which acts through Or47b-expressing ORNs, our findings indicate that 7-T and ppk23 neurons, which are capable of detecting 7-T, also play a role in MIES induction. In D. melanogaster and other related species, food odors typically serve as volatile long-range signals that attract both males and females (74, 75), suggesting that specific food odors may also influence EHP (42). The involvement of multiple ORNs in the regulation of EHP predicts that pC1 neurons may process multiple odorants, not limited to those associated with mating behavior, including food odors. Future studies will explore the full spectrum of odorants processed by pC1 neurons in the regulation of EHP.

In conclusion, we have identified a circuit that, via the detection of a novel male pheromone, potentially signals male quality and governs the female’s decision to remove the mating plug of her last mate and mate again.

Materials and Methods

Fly care

Flies were cultured on a standard medium composed of dextrose, corn meal, and yeast, at room temperature on a 12hr : 12hr light:dark cycle (40, 73). Behavioral assays were performed at 25 °C, except for the thermogenetic activation experiment with dTRPA1. Virgin males and females were collected immediately after eclosion. Males were aged individually for 4-6 days, while females were aged in groups of 15–20. For EHP and mating assays, females were aged for 3-4 days. Assays were performed at Zeitgeber time (ZT) 3:00–11:00 and were repeated on at least three separate days.

Fly stocks

The following stocks are from the Bloomington Drosophila Stock Center (BDSC), the Vienna Drosophila Resource Center (VDRC): Canton S (CS) (RRID: BDSC_64349), w1118 (VDRC #60000), R71G01 (pC1-Gal4) (RRID: BDSC_39599), Orco1 (RRID: BDSC_23129), Or13a-Gal4 (RRID: BDSC_9946), Or19a-Gal4 (RRID: BDSC_9948), Or23a-Gal4 (RRID: BDSC_9955), Or43a-Gal4 (RRID: BDSC_9974), Or47b-Gal4 (RRID: BDSC_9983), Or47b-Gal4 (RRID: BDSC_9984), Or65a-Gal4 (RRID: BDSC_9993), Or65b-Gal4 (RRID: BDSC_23901), Or65c-Gal4 (RRID: BDSC_23903), Or67d-Gal4 (RRID: BDSC_9998), Or83c-Gal4 (RRID: BDSC_23131), Or88a-Gal4 (RRID: BDSC_23137), UAS-Or47b (RRID: BDSC_76045), Or47b2/2 (RRID: BDSC_51306), Or47b3/3 (RRID: BDSC_51307), UAS-TNT active (RRID: BDSC_28837), UAS-TNT inactive (RRID: BDSC_28839), UAS-dTRPA1 (RRID: BDSC_26263), UAS-CsChrimson (RRID: BDSC_55135), uAS-GCaMP6m (RRID: BDSC_42748), R52G04-AD (RRID: BDSC_71085), SAG-Gal4 (VT50405) (RRID:Flybase_FBst0489354, VDRC #200652), UAS-Dh44R1-RNAi (RRID:Flybase_FBst0482273, VDRC #110708), UAS-Dh44R2-RNAi (RRID:Flybase_FBst0465025, VDRC #43314), UAS-Dicer2 (VDRC #60007). The following stocks were previously reported: PromE(800)-Gal4 (59), UAS-FLP, CRE-F-luc (76), LexAop-FLP (77), UAS-CsChrimson (78), UAS-GtACR1 (79), UAS-PhotoAC (PACα) (80), pC1-A (48), pC1-S (48), Dh44-pC1-Gal4 (47), ppk23-Gal4, ppk23-, ppk28-, ppk29-(62), and Orco-Gal4, UAS-EGFP-Orco (81). pC1a-split-Gal4 is generated by combining R52G04-AD (RRID: BDSC_71085) and dsx-DBD (49). Drosophila species other than D. melanogaster are obtained from the EHIME-Fly Drosophila Stock Center and the KYORIN-Fly Drosophila species Stock Center. To enhance knock-down efficiency, RNAi experiments were performed using flies carrying UAS-Dicer2 (VDRC #60007).

Chemical information

All trans-retinal (Cat# R2500), methyl laurate (Cat# W271500), and Triton™ X-100 (Cat# X100) were obtained from Sigma-Aldrich (St. Louis, MO, USA). The following chemicals were obtained from the Cayman Chemical (Ann Arbor, MI, USA): 7(Z)-Tricosene (CAS No. 52078-42-9, Cat# 9000313), 7(Z)-Pentacosene (CAS No. 63623-49-4, Cat# 9000530), trans-palmitoleic acid (CAS No. 10030-73-6, Cat# 9001798), 11-cis-vaccenyl acetate (cVA) dissolved in EtOH (CAS No. 6186-98-7, Cat# 10010101). 2-methyltetracosane (>98% purity) was custom-synthesized by KIP (Daejeon, Korea). Ethanol is used as a vehicle for 7-T, cVA, trans-palmitoleic acid and methyl laurate, while hexane is used as a vehicle for 2-methyltetracosane.

Behavior assays

For mating behavior assays, we followed the procedures described previously (82). Individual virgin females and naive CS males were paired in 10 mm diameter chambers and were recorded using a digital camcorder (SONY, HDR-CX405 or Xiaomi, Redmi Note 10) for either 30 minutes or 1 hour for the mating assay and 6 hours for the re-mating assay. In the re-mating assay, females that completed their initial mating within 30 minutes were subsequently paired with naive CS males.

To measure EHP, defined as the time elapsed between the end of copulation and sperm ejection, we used the following procedure: Virgin females were individually mated with CS males in 10 mm diameter chambers. Following copulation, females were transferred to new chambers, either with or without a CS male or pheromone presentation, and their behavior was recorded using a digital camcorder (SONY, HDR-CX405). Typically, females that completed copulation within 30 minutes were used for analysis. The sperm ejection scene, in which the female expels a white sac containing sperm and the mating plug through the vulva, was directly observed by eye in the recorded video footage. For pheromone presentation, females were individually housed in 10 mm diameter chambers containing a piece of Whatman filter paper (2 mm x 2 mm) treated with 0.5 μl of the pheromone solution and air dried for 1 minute. For thermogenetic activation experiments, females were incubated at the indicated temperatures immediately after the end of copulation. For light activation experiments, a custom-made light activation setup was used with a ring of 104 multi-channel LED lights (NeoPixel, Cat# WS2812; red light, 620-625 nm, 390-420 mcd; green light, 522-525 nm, 660-720 mcd; blue light, 465-467 nm, 180-200 mcd). Females were individually placed in 10 mm diameter chambers, and the chamber was illuminated with light at an intensity of 1100 lux across the chamber during the assay, as measured by an HS1010 light meter. Flies used in these experiments were prepared by culturing them immediately after eclosion in food containing vehicle (EtOH) or 1 mM all trans-retinal (ATR). They were kept in complete darkness for 3-4 days until the assay was conducted. To prevent the accumulation of residual pheromones, all behavioral chambers were cleaned with 70% water/ethanol or acetone before and after the experiment.

Calcium imaging

We followed the procedures described previously (47, 83). Following copulation, freshly mated female flies were temporarily immobilized using ice anesthesia, and their heads were attached to a custom-made thin metal plate with a 1-mm diameter hole using photo-curable UV glue (ThreeBond, A16A01). An opening in the fly’s head was created using a syringe needle under saline (108 mM NaCl, 5 mM KCl, 2 mM CaCl2, 8.2 mM MgCl2, 4 mM NaHCO3, 1 mM NaH2PO4, 5 mM trehalose, 10 mM sucrose, 5 mM HEPES pH 7.5). Imaging was performed with a Zeiss Axio Examiner A1 microscope equipped with an electron-multiplying CCD camera (Andor Technology, LucaEM R 604M) and an LED light source (CoolLED, Precis Excite). Metamorph software (Molecular Devices, RRID:SCR_002368) was used for image analysis. The Syntech Stimulus Controller (Type CS-55) was used to deliver the male pheromone using an airflow. 2 μl of pheromone solution was applied to a piece of Whatman filter paper (2 mm x 1 mm), which was then inserted into a glass Pasteur pipette after solvent evaporation.

Luciferase assay

We followed the procedures described previously (47, 76). For the assay, 3-day-old virgin females or freshly mated females were used. A group of three fly heads, kept at -80°C, was homogenized using cold homogenization buffer (15 mM HEPES, 10 mM KCl, 5 mM MgCl2, 0.1 mM EDTA, 0.5 mM EGTA).

Luciferase activity was measured using beetle luciferin potassium salt (Promega, Cat# E1603) and a microplate luminometer (Berthold technologies, Centro XS3 LB 960), following the manufacturer’s instructions. For pheromone presentation, flies were placed in 10 mm diameter chambers containing a piece of Whatman filter paper (4 mm x 6 mm) treated with 1 μl of the pheromone solution and air dried for 1 minute.

Immunohistochemistry

3–5-day-old virgin female flies were dissected in phosphate buffered saline (PBS) and fixed in 4% paraformaldehyde in PBS for 30 minutes at room temperature. After fixation, the brains were thoroughly washed in PBST (0.1% Triton™ X-100 in PBS) and then blocked with 5% normal goat serum in PBST. After blocking, brains were incubated with primary antibody in PBST for 48 hours at 4°C, washed with PBST, and then incubated with secondary antibody in PBST for 24 hours at 4°C. The samples were washed three times with PBST and once with PBS before mounting in Vectashield (Vector Laboratories, Cat# H-1000). Antibodies used were rabbit anti-GFP (1:1000; Thermo Fisher Scientific, Cat# A-11122, RRID:AB_221569), mouse anti-nc82 (1:50; Developmental Studies Hybridoma Bank, Cat# Nc82; RRID: AB_2314866), Alexa 488-conjugated goat anti-rabbit (1:1000; Thermo Fisher Scientific, Cat# A-11008, RRID:AB_143165), Alexa 568-conjugated goat anti-mouse (1:1000; Thermo Fisher Scientific, Cat# A-11004, RRID:AB_2534072). Brain images were acquired with a Zeiss LSM 700/Axiovert 200M (Zeiss) and processed with Fiji (https://imagej.net/software/fiji/downloads, RRID:SCR_002285)

Color depth MIP based anatomical analysis

A stack of confocal images of pC1a-split-Gal4>UAS-myr-EGFP adult female brains stained with anti-GFP and anti-nc82 was used. Images were registered to the JRC2018 unisex brain template (84) using the Computational Morphometry Toolkit (CMTK, https://github.com/jefferis/fiji-cmtk-gui). Color depth MIP masks of pC1a-split-Gal4 neurons and pC1a (ID, 5813046951) in Hemibrain (72) (Fig. S6C) were generated using the ColorMIP_Mask_Search plugin (85) for Fiji (https://github.com/JaneliaSciComp/ColorMIP_Mask_Search) and NeuronBridge (86) (https://neuronbridge.janelia.org/). Similarity score and rank were calculated using NeuronBridge.

Statistical analysis

Statistical analysis was conducted using GraphPad Prism 9 (Graphpad, RRID:SCR_002798), with specific details of each statistical method provided in the figure legends.

Supporting information